---
title: "Spatial analyses of excess mortality in CH"
subtitle: "Geodata preparation"
author: "Radek Panczak"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
    code_folding: show
    self_contained: true
    highlight: pygments
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../docs") })
---

<!-- ----------------------------------------------------- -->

```{r setup, include = FALSE}
set.seed(12345)
options(scipen = 999)
options(max.print = "75")

library(pacman)
p_load(sf, tidyverse, readxl, magrittr, 
       scales, janitor, lubridate)

import::from("sjmisc", "frq")
```


```{r setup, include = FALSE}
tidyverse::tidyverse_conflicts()

conflicted::conflict_scout()
```

```{r knit-setup, include = FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

knitr::opts_chunk$set(cache = FALSE,
                      prompt = FALSE,
                      tidy = FALSE,
                      comment = NA,
                      message = FALSE,
                      warning = FALSE)

knitr::opts_knit$set(width = 75)

mu <- Hmisc::markupSpecs$html
```

<!-- ----------------------------------------------------- -->

# Data sources 

## BfS data 2021

[Generalisierte Gemeindegrenzen: Geodaten](https://www.bfs.admin.ch/bfs/de/home/dienstleistungen/geostat/geodaten-bundesstatistik/administrative-grenzen/generalisierte-gemeindegrenzen.assetdetail.17964056.html).  

Contains three versions of data:  

> "Generalisierte Gemeindegrenzen, Stand 01.01.2021 & 18.04.2021 & 01.07.2021"	

Currently using `01.07` edition.  

```{r eval=FALSE}
download.file(url = "https://www.bfs.admin.ch/bfsstatic/dam/assets/17964056/master",
              destfile = "data-raw/BfS/ag-b-00.03-875-gg21.zip",
              method = "curl")

unzip("data-raw/BfS/ag-b-00.03-875-gg21.zip", exdir = "data-raw/BfS/ag-b-00.03-875-gg21")

unlink("data-raw/BfS/ag-b-00.03-875-gg21.zip")
unlink("data-raw/BfS/ag-b-00.03-875-gg21/txt", recursive = TRUE)
unlink("data-raw/BfS/ag-b-00.03-875-gg21/kmz", recursive = TRUE)
unlink("data-raw/BfS/ag-b-00.03-875-gg21/ggg_2021-LV03", recursive = TRUE)
unlink("data-raw/BfS/ag-b-00.03-875-gg21/ggg_2021-LV95/ggg_2021-LV95.gdb", recursive = TRUE)
```

```{r}
gg21 <- 
  st_read("data-raw/BfS/ag-b-00.03-875-gg21/ggg_2021-LV95/shp/g1g21_01072021.shp") 

bz21 <- 
  st_read("data-raw/BfS/ag-b-00.03-875-gg21/ggg_2021-LV95/shp/g1b21.shp")

kt21 <- 
  st_read("data-raw/BfS/ag-b-00.03-875-gg21/ggg_2021-LV95/shp/g1k21.shp")

se21 <- 
  st_read("data-raw/BfS/ag-b-00.03-875-gg21/ggg_2021-LV95/shp/g1s21.shp")

write_rds(bz21, "data/BfS/bz21.Rds")
write_rds(kt21, "data/BfS/kt21.Rds")
write_rds(se21, "data/BfS/se21.Rds")
```

```{r eval=FALSE, include=FALSE}
kt21 %>% 
  st_geometry() %>% 
  plot()
```

## swisstopo data 2021

`swissBOUNDARIES3D` data available from [here](https://www.swisstopo.admin.ch/de/geodata/landscape/boundaries3d.html). Also exist for 1st Jan and XX July versions.  

```{r eval=FALSE}
download.file(url = "https://data.geo.admin.ch/ch.swisstopo.swissboundaries3d/swissboundaries3d_2021-07/swissboundaries3d_2021-07_2056_5728.shp.zip",
              destfile = "data-raw/swisstopo/swissboundaries3d_2021-07_2056_5728.shp.zip",
              method = "curl")

unzip("data-raw/swisstopo/swissboundaries3d_2021-07_2056_5728.shp.zip", 
      exdir = "data-raw/swisstopo/swissboundaries3d_2021-07_2056_5728")

unlink("data-raw/swisstopo/swissboundaries3d_2021-01_2056_5728.shp.zip")
```

```{r}
st_gg21 <- 
  st_read("data-raw/swisstopo/swissboundaries3d_2021-07_2056_5728/SHAPEFILE_LV95_LN02/swissBOUNDARIES3D_1_3_TLM_HOHEITSGEBIET.shp") %>% 
  st_zm(drop = TRUE, what = "ZM")

# st_bz21 <- 
#   st_read("data-raw/swisstopo/swissboundaries3d_2021-07_2056_5728/SHAPEFILE_LV95_LN02/swissBOUNDARIES3D_1_3_TLM_BEZIRKSGEBIET.shp") %>% 
# st_zm(drop = TRUE, what = "ZM")

st_kt21 <- 
  st_read("data-raw/swisstopo/swissboundaries3d_2021-07_2056_5728/SHAPEFILE_LV95_LN02/swissBOUNDARIES3D_1_3_TLM_KANTONSGEBIET.shp") %>% 
  st_zm(drop = TRUE, what = "ZM")

# write_rds(st_bz21, "data/swisstopo/st_bz21.Rds")
write_rds(st_kt21, "data/swisstopo/st_kt21.Rds")
```


# Preparations

## Non-residential communities

Excluding `Staatswald Galm`, `Comunanza Cadenazzo/Monteceneri` & `Comunanza Capriasca/Lugano`.  

```{r}
gg21 %>% 
  st_drop_geometry() %>% 
  filter(GMDNR %in% c(2391, 5391, 5394)) %>% 
  select(GMDNAME)
```

```{r include=FALSE}
gg21 %<>% 
  filter(!GMDNR %in% c(2391, 5391, 5394)) %>% 
  select(GMDNR, GMDNAME, BZNR, KTNR) %>% 
  arrange(GMDNR)

st_gg21 %<>% 
  filter(!BFS_NUMMER %in% c(2391, 5391, 5394)) %>% 
  filter(OBJEKTART != "Kantonsgebiet") %>% 
  filter(ICC == "CH") %>% 
  select(BFS_NUMMER, NAME, BEZIRKSNUM, KANTONSNUM, GEM_TEIL) %>% 
  rename(GMDNR = BFS_NUMMER,
         GMDNAME = NAME,
         BZNR = BEZIRKSNUM,
         KTNR = KANTONSNUM) %>% 
  arrange(GMDNR)
```

## Multipart features

Example of Monthey 

`gg21` is `MULTIPOLYGON`

`st_gg21` is `POLYGON`


```{r}
gg21 %>% 
  filter(GMDNR == 6153) %>% 
  select(GMDNR) %>% 
  plot()

st_gg21 %>% 
  filter(GMDNR == 6153) %>%
  select(GEM_TEIL) %>% 
  plot()
```


```{r eval=FALSE}
gg21 %>%
  filter(! GMDNR %in% st_gg21$BFS_NUMMER) %>% 
  nrow()

st_gg21 %>%
  filter(! BFS_NUMMER %in% gg21$GMDNR) %>% 
  nrow()
```


```{r}
write_rds(gg21, "data/BfS/gg21.Rds")
write_rds(st_gg21, "data/swisstopo/st_gg21.Rds")
```


# Community boundaries mutations

Using `Gemeindestand - Stand vom 01.07.2021.xlsx` from [here](https://www.agvchapp.bfs.admin.ch/de/home) we can identify communities with changes during the study period.  

```{r}
histcomm <- read_excel("data-raw/BfS/Gemeindestand - Stand vom 01.07.2021.xlsx", 
                       col_types = c("numeric", "text", "skip", 
                                     "skip", "numeric", "text", "text")) %>%
  remove_empty(c("rows", "cols")) %>% clean_names() %>% 
  mutate(datum_der_aufnahme = ymd(datum_der_aufnahme)) %>% 
  filter(datum_der_aufnahme >= ymd("2015-01-01") & 
           datum_der_aufnahme < ymd("2021-07-01")) %>% 
  arrange(desc(datum_der_aufnahme), gemeindename)
```

There were `r length(unique(histcomm$bfs_gde_nummer))` communities with changes recorded:  

```{r echo=FALSE}
histcomm %>% 
  kbl() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>% 
  column_spec(1, italic = TRUE) %>%
  row_spec(1, italic = FALSE)
```


```{r eval=FALSE, include=FALSE}
# Auf den 18. April 2021 fusionierten die ehemaligen politischen Gemeinden Croglio, Monteggio, Ponte Tresa und Sessa zur neuen politischen Gemeinde Tresa.

histcomm %>% 
  filter(bfs_gde_nummer == 5239)

histcomm %>% 
  filter(bfs_gde_nummer %in% c(5178, 5202, 5213, 5222))

gg21 %>% 
  st_drop_geometry() %>% 
  filter(GMDNR %in% c(5178, 5202, 5213, 5222, 5239)) %>% 
  select(GMDNR, GMDNAME)

st_read("data-raw/BfS/ag-b-00.03-875-gg21/ggg_2021-LV95/shp/g1g21_01012021.shp",
        quiet = TRUE) %>% 
  st_drop_geometry() %>% 
  filter(GMDNR %in% c(5178, 5202, 5213, 5222, 5239)) %>% 
  select(GMDNR, GMDNAME)
```


