---
title: "Spatial analyses of 2020-21 excess mortality in CH"
subtitle: "Mortality data preparation (*Wohnort* based)"
author: "Radek Panczak"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
    code_folding: show
    self_contained: true
    highlight: pygments
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../docs") })
---

<!-- ----------------------------------------------------- -->

```{r setup, include = FALSE}
set.seed(12345)
options(scipen = 999)
options(max.print = "75")

library(pacman)
p_load(sf, tidyverse, magrittr, janitor, scales, lubridate, DT)

import::from("sjmisc", "frq")
```

```{r conflicts, include = FALSE}
tidyverse::tidyverse_conflicts()

conflicted::conflict_scout()
```

```{r knit-setup, include = FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

knitr::opts_chunk$set(cache = FALSE,
                      prompt = FALSE,
                      tidy = FALSE,
                      comment = NA,
                      message = FALSE,
                      warning = FALSE)

knitr::opts_knit$set(width = 75)

mu <- Hmisc::markupSpecs$html
```

<!-- ----------------------------------------------------- -->

# Mortality data 2015-2021

## Source 

Custom file delivered by BfS covering 2021 period and standardizing communities to Jul 2021.  

Using **Wohnort** (the community where the person died) as unit of anlysis.   

```{r}
w_deaths_2015_2021 <- read_csv("data-raw/BfS-closed/13OCT21_Lieferung_Panczak.csv", 
                             col_types = cols(year = col_integer(), 
                                              month = col_integer(), 
                                              COUNT = col_integer()),
                             locale = readr::locale(encoding = "latin1")) %>% 
  mutate(sex = if_else(sex == "F", "Female", "Male"),
         nationality = if_else(nationality == "Schweiz", "Swiss", "Foreigner")) %>% 
  rename(GMDNAME = community) %>% 
  # hyphen in name >> doesnt link to spatial data
  mutate(GMDNAME = if_else(GMDNAME == "La Grande-Béroche", "La Grande Béroche", GMDNAME)) %>% 
  # name change? >> TBresolved with Cordula
  mutate(GMDNAME = if_else(GMDNAME == "Klosters-Serneus", "Klosters", GMDNAME)) %>% 
  group_by(year, month, canton, GMDNAME, agegroup, sex, nationality) %>% 
  summarise(COUNT = sum(COUNT)) %>% 
  ungroup()
```

Description of data:  

```{r echo=FALSE}
report::report(w_deaths_2015_2021)
```

```{r eval=FALSE, include=FALSE}
# Deaths across variables:  
w_deaths_2015_2021 %>%
  mutate(month = factor(month)) %>% 
  uncount(deaths) %>% 
  select(-GMDNAME, -canton) %>% 
  gtsummary::tbl_summary(by = year)
```

```{r eval=FALSE, include=FALSE}
# Canton x year distribution:  
w_deaths_2015_2021 %>%
  uncount(deaths) %>% 
  select(year, canton) %>% 
  gtsummary::tbl_summary(by = year)
```

## Codes of geographical areas

```{r include=FALSE}
gg21 <- read_rds("data/BfS/gg21.Rds") %>% 
  st_drop_geometry() %>% 
  # prepared in 01.Rmd
  left_join(read_rds("data/BfS/AR.Rds"))
```

```{r include=FALSE}
# Codes with no link
# has to be empty 

no_link <- w_deaths_2015_2021 %>% 
  select(GMDNAME) %>% 
  distinct()  %>% 
  anti_join(gg21)

stopifnot(nrow(no_link) == 0)

rm(no_link)
```

There were `r number(nrow(gg21), big.mark = ",")` communities in Jul 2021 in CH.  

```{r echo=FALSE}
no_cases <- gg21 %>% 
  anti_join(
    w_deaths_2015_2021 %>% 
      select(GMDNAME) %>% 
      distinct() 
  )
```

`r number(nrow(no_cases), big.mark = ",")` had no deaths recorded during the study period.   

```{r echo=FALSE}
no_cases %>% 
  select(GMDNAME, KTNAME) %>% 
  datatable(options = list(paging = FALSE)) 

rm(no_cases)
```

## Age groups

Original age groups:  

```{r echo=FALSE}
frq(w_deaths_2015_2021, agegroup)
```

Were simplified to `<40`, `40-59` , `60-69`, `70-79` & `80+`

```{r}
w_deaths_2015_2021 %<>% 
  mutate(
    age = case_when(
      agegroup == "[0-10)" ~   "<40",
      agegroup == "[10-20)" ~  "<40",
      agegroup == "[20-30)" ~  "<40",
      agegroup == "[30-40)" ~  "<40",
      
      agegroup == "[40-50)" ~  "40-49",
      agegroup == "[50-60)" ~  "50-59",
      agegroup == "[60-70)" ~  "60-69",
      agegroup == "[70-80)" ~  "70-79",
      
      agegroup == "[80-90)" ~  "80+",
      agegroup == "[90-116)" ~ "80+" ,
      TRUE ~                    agegroup)
  ) %>% 
  select(-agegroup) %>% 
  group_by(year, month, canton, GMDNAME, sex, age, nationality) %>% 
  summarise(deaths = sum(COUNT)) %>% 
  ungroup()
```

```{r echo=FALSE}
frq(w_deaths_2015_2021, age)
```

## Distribution of death counts

Obviously highly skewed.  
Note:  No zeroes here yet! No pop denominator either!  

```{r}
w_deaths_2015_2021 %>% 
  ggplot(aes(deaths)) + 
  geom_histogram(binwidth = 1)
```

## Filling with zeroes

All possible combinations of `year, month, canton, community, age, sex, nationality` are used and then empty strata are filled with zeroes.  

This expansion also includes communities with zero deaths described above.  

```{r}
w_deaths_2015_2021_exp <- 
  gg21 %>% 
  select(GMDNAME) %>% 
  distinct() %>% 
  left_join(w_deaths_2015_2021) %>% 
  select(-deaths, -canton) %>% 
  # avoiding NAs
  mutate(
    year = if_else(is.na(year), as.integer(2015), year),
    month = if_else(is.na(month), as.integer(1), month),
    sex = if_else(is.na(sex), "Male", sex),
    age = if_else(is.na(age), "<40", age),
    nationality = if_else(is.na(nationality), "Swiss", nationality),
  ) %>% 
  expand(GMDNAME, year, month, age, sex, nationality) %>% 
  left_join(w_deaths_2015_2021) %>% 
  replace_na(list(deaths = 0)) %>% 
  mutate(date = ymd(paste(year, month, "1", sep = " "))) %>% 
  relocate(date, .after = month) %>% 
  relocate(canton, .after = GMDNAME) %>% 
  group_by(GMDNAME) %>% 
  fill(canton, .direction = "downup") %>% 
  ungroup()
```

That obviously increases the size of the dataset from `r number(nrow(w_deaths_2015_2021), big.mark = ",")` to `r number(nrow(w_deaths_2015_2021_exp), big.mark = ",")` (*ie.* 2,163 communities each with 2,016 data points).  

Needless to say that creates large amount of zero cells:

```{r echo=FALSE}
frq(w_deaths_2015_2021_exp, deaths == 0)
```

## Examples of time series

### Communities Bern & Meienried:  

```{r echo=FALSE}
w_deaths_2015_2021_exp %>% 
  filter(GMDNAME %in% c("Bern", "Meienried")) %>% 
  group_by(GMDNAME, date) %>% 
  summarise(deaths = sum(deaths)) %>% 
  ungroup() %>% 
  ggplot(aes(x = date, y = deaths, group = GMDNAME, color = GMDNAME)) + 
  geom_line() + geom_point()
```

## All time series

```{r echo=FALSE}
w_deaths_2015_2021_exp %>% 
  group_by(GMDNAME, date) %>% 
  summarise(deaths = sum(deaths)) %>% 
  ungroup() %>% 
  ggplot(aes(x = date, y = deaths, group = GMDNAME)) + 
  geom_line(colour = "darkorchid", alpha = 0.25) +
  theme_minimal()
```

```{r include=FALSE}
# write_rds(w_deaths_2015_2021, "data/BfS-closed/w_deaths_2015_2021.Rds")
write_rds(w_deaths_2015_2021_exp, "data/BfS-closed/w_deaths_2015_2021_exp.Rds")
```

<!-- ----------------------------------------------------- -->

# Simplification of data

## Aggregating to districts

```{r}
w_deaths_2015_2021_bez <- w_deaths_2015_2021_exp %>% 
  left_join(gg21 %>% select(GMDNAME, BZNR)) %>% 
  select(-GMDNAME) %>% 
  group_by(BZNR, canton, year, month, date, age, sex, nationality) %>% 
  summarise(deaths = as.integer(sum(deaths))) %>% 
  ungroup()
```

That reduces the size of the dataset to `r number(nrow(w_deaths_2015_2021_bez), big.mark = ",")`.  

Proportion of zero cells decreases:

```{r echo=FALSE}
frq(w_deaths_2015_2021_bez, deaths == 0)
```

## Examples of time series

### Districts Bern & Seeland

```{r echo=FALSE}
w_deaths_2015_2021_bez %>% 
  filter(BZNR %in% c(246, 243)) %>% 
  group_by(BZNR, date) %>% 
  summarise(deaths = sum(deaths)) %>% 
  ungroup() %>% 
  ggplot(aes(x = date, y = deaths, group = factor(BZNR), color = factor(BZNR))) + 
  geom_line() + geom_point()
```

## All time series

```{r echo=FALSE}
w_deaths_2015_2021_bez %>% 
  group_by(BZNR, date) %>% 
  summarise(deaths = sum(deaths)) %>% 
  ungroup() %>% 
  ggplot(aes(x = date, y = deaths, group = BZNR)) + 
  geom_line(colour = "darkorchid", alpha = 0.25) +
  theme_minimal()
```

```{r include=FALSE}
write_rds(w_deaths_2015_2021_bez, "data/BfS-closed/w_deaths_2015_2021_bez.Rds")
```

## Aggregating to Arbeitsmarktgrossregionen 

```{r}
w_deaths_2015_2021_ar <- w_deaths_2015_2021_exp %>% 
  left_join(gg21 %>% select(GMDNAME, ARGRNR, ARGRNAME, ARNR, ARNAME)) %>% 
  select(-GMDNAME) %>% 
  group_by(ARGRNR, ARGRNAME, ARNR, ARNAME, 
           canton, year, month, date, age, sex, nationality) %>% 
  summarise(deaths = as.integer(sum(deaths))) %>% 
  ungroup()
```

That reduces the size of the dataset to `r number(nrow(w_deaths_2015_2021_ar), big.mark = ",")`.  

Proportion of zero cells:

```{r echo=FALSE}
frq(w_deaths_2015_2021_ar, deaths == 0)
```

## Examples of time series

### Districts Bern & Biel/Bienne

```{r echo=FALSE}
w_deaths_2015_2021_ar %>% 
  filter(ARNAME %in% c("Bern", "Biel/Bienne")) %>% 
  group_by(ARNAME, date) %>% 
  summarise(deaths = sum(deaths)) %>% 
  ungroup() %>% 
  ggplot(aes(x = date, y = deaths, group = factor(ARNAME), color = factor(ARNAME))) + 
  geom_line() + geom_point()
```

## All time series

```{r echo=FALSE}
w_deaths_2015_2021_ar %>% 
  group_by(ARNAME, date) %>% 
  summarise(deaths = sum(deaths)) %>% 
  ungroup() %>% 
  ggplot(aes(x = date, y = deaths, group = ARNAME)) + 
  geom_line(colour = "darkorchid", alpha = 0.25) +
  theme_minimal()
```

```{r include=FALSE}
write_rds(w_deaths_2015_2021_ar, "data/BfS-closed/w_deaths_2015_2021_ar.Rds")
```