---
title: "Spatial analyses of 2020-21 excess mortality in CH"
subtitle: "Ancillary data preparation"
author: "Radek Panczak"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
    code_folding: show
    self_contained: true
    highlight: pygments
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../docs") })
---

<!-- ----------------------------------------------------- -->

```{r setup, include = FALSE}
set.seed(12345)
options(scipen = 999)
options(max.print = "75")

library(pacman)
p_load(haven, readxl,
       tidyverse, magrittr, janitor, scales, lubridate, 
       DT, 
       sf, tmap)

import::from("sjmisc", "frq")
```


```{r conflicts, include = FALSE}
tidyverse::tidyverse_conflicts()

conflicted::conflict_scout()
```

```{r knit-setup, include = FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

knitr::opts_chunk$set(cache = FALSE,
                      prompt = FALSE,
                      tidy = FALSE,
                      comment = NA,
                      message = FALSE,
                      warning = FALSE)

knitr::opts_knit$set(width = 75)

mu <- Hmisc::markupSpecs$html
```

<!-- ----------------------------------------------------- -->
<!-- ----------------------------------------------------- -->

# Population

Data extracted from [`px-x-0103010000_201`](https://www.pxweb.bfs.admin.ch/pxweb/de/px-x-0103010000_201/-/px-x-0103010000_201.px/):

> Ständige und nichtständige Wohnbevölkerung nach institutionellen Gliederungen, Staatsangehörigkeit (Kategorie), Geburtsort, Geschlecht und Altersklasse

Important info from footnotes:

> *Letzte Änderungen: Neuer Datensatz (Jahr 2020)*
> *Stand der Datenbank: Juni 2021*
> *Stichtag: 31. Dezember*
> **Raumbezug: Gemeinden / 18.10.2020**
> *Datenquelle: Statistik der Bevölkerung und der Haushalte STATPOP*
> *Definition der ständigen Wohnbevölkerung*

## Community codes

Community mutations data after `2021-01-01` from @Cordula are integrated into the data and pops are recalculated using these new codes to match mortality.  

```{r}
histcomm <- read_rds("data/BfS/histcomm.Rds") %>% 
  filter(datum_der_aufnahme >= ymd("2021-01-01")) %>% 
  select(bfs_gde_num_old, bfs_gde_num_new, gemeindename_new)
```

```{r eval=FALSE, include=FALSE}
datatable(histcomm)
```

```{r eval=FALSE, include=FALSE}
histcomm_14 <- read_excel("data-raw/BfS/Gemeindestand - Stand vom 01.07.2021.xlsx", 
                          col_types = c("numeric", "text", "skip", 
                                        "skip", "numeric", "text", "text")) %>%
  remove_empty(c("rows", "cols")) %>% clean_names() %>% 
  mutate(datum_der_aufnahme = ymd(datum_der_aufnahme)) %>% 
  filter(datum_der_aufnahme >= ymd("2014-01-01") & 
           datum_der_aufnahme < ymd("2015-01-01")) %>% 
  arrange(desc(datum_der_aufnahme), gemeindename)
```

## Data preps

Data aggregated to age groups as in deaths dataset. 

```{r}
population <- read_xlsx("data-raw/BfS/px-x-0103010000_201.xlsx", 
                        col_types = c("numeric", 
                                      "skip", "numeric", "text", "skip", 
                                      "skip", "skip", "text", "skip", "skip", 
                                      "skip", "text", "numeric", "numeric", 
                                      "numeric", "numeric", "numeric", 
                                      "numeric", "numeric", "numeric", 
                                      "numeric", "numeric", "numeric", 
                                      "numeric", "numeric", "numeric", 
                                      "numeric", "numeric", "numeric", 
                                      "numeric", "numeric", "numeric", 
                                      "numeric"), 
                        skip = 1) %>%
  remove_empty(c("rows", "cols")) %>% clean_names() %>% 
  rename(year = x1, 
         GMDNR = x2,
         GMDNAME = x3) %>% 
  mutate(sex = if_else(x5 == "Frau", "Female", "Male"),
         nationality = if_else(x4 == "Schweiz", "Swiss", "Foreigner")) %>% 
  select(-x4, -x5) %>% 
  relocate(sex, .after = GMDNAME) %>% 
  relocate(nationality, .after = sex) %>% 
  mutate(GMDNAME = word(GMDNAME, 2, -1)) %>% 
  fill(year, GMDNR, GMDNAME, nationality) %>% 
  mutate(`<40` = x0_4_jahre + x5_9_jahre + x10_14_jahre + x15_19_jahre + 
           x20_24_jahre + x25_29_jahre + x30_34_jahre + x35_39_jahre) %>% 
  select(-x0_4_jahre, -x5_9_jahre, -x10_14_jahre, -x15_19_jahre,  
         -x20_24_jahre, -x25_29_jahre, -x30_34_jahre, -x35_39_jahre) %>% 
  mutate(`40-49` = x40_44_jahre + x45_49_jahre) %>% 
  mutate(`50-59` = x50_54_jahre + x55_59_jahre) %>% 
  mutate(`60-69` = x60_64_jahre + x65_69_jahre) %>% 
  mutate(`70-79` = x70_74_jahre + x75_79_jahre) %>% 
  select(-x40_44_jahre, -x45_49_jahre, -x50_54_jahre, -x55_59_jahre,
         -x60_64_jahre, -x65_69_jahre, -x70_74_jahre, -x75_79_jahre) %>% 
  mutate(`80+` = x80_84_jahre +x85_89_jahre + x90_94_jahre +
           x95_99_jahre + x100_jahre_und_mehr) %>% 
  select(-x80_84_jahre, -x85_89_jahre, -x90_94_jahre,
         -x95_99_jahre, -x100_jahre_und_mehr) %>% 
  pivot_longer(cols = `<40`:`80+`, 
               names_to = "age", values_to = "population") %>% 
  left_join(histcomm, by = c("GMDNR" = "bfs_gde_num_old")) %>% 
  mutate(GMDNR = if_else(!is.na(bfs_gde_num_new), bfs_gde_num_new, GMDNR),
         GMDNAME = if_else(!is.na(gemeindename_new), gemeindename_new, GMDNAME)) %>% 
  select(-bfs_gde_num_new, -gemeindename_new) %>% 
  group_by(year, GMDNR, GMDNAME, sex, nationality, age) %>% 
  summarise(population = sum(population)) %>% 
  ungroup() %>% 
  arrange(year, GMDNR, sex, nationality, age)
```

Yearly totals (**on 31st Dec**!):  

```{r echo=FALSE}
population %>% 
  group_by(year) %>% 
  summarise(population = sum(population)) %>% 
  mutate(population = as.character(number(population, big.mark = ","))) %>% 
  datatable(rownames = FALSE, options = list(dom = 't'))
```

Age (40 plus!) distributions over years  

```{r echo=FALSE}
population %>% 
  filter(age != "<40") %>% 
  group_by(year, age, sex) %>% 
  summarise(population = sum(population)) %>% 
  ungroup() %>% 
  ggplot(aes(x = age, y = population, fill = as.factor(year))) + 
  geom_col(position = "dodge") + 
  scale_fill_viridis_d()
```

```{r include=FALSE}
write_rds(population, "data/BfS/population.Rds")
```

## Predicting 2020 & 2021 pops

```{r eval=FALSE, include=FALSE}
# smallest pops
population %>% 
  filter(year == 2020) %>% 
  group_by(GMDNR) %>% 
  summarize(population = sum(population),
            GMDNAME = first(GMDNAME)) %>% 
  ungroup() %>% 
  arrange(population)
# arrange(desc(population))
```

For each age, sex, nationality stratum, for each municipality separately, ~~poisson~~ linear model is fit using the 2014-2019 data; then prediction is made for 2020 & 2021 years. The gist of this operation is to exclude the effect of pandemic from the population counts in years affected.   

```{r}
population_ext_lm <- population %>% 
  filter(year != 2020) %>% 
  # filter(GMDNR == 261 | GMDNR == 782) %>%
  group_by(GMDNR, age, sex, nationality) %>% 
  # do(broom::tidy(lm(population ~ year, data = .))) %>% 
  # do(broom::augment(lm(population ~ year, data = .))) %>% 
  do(lm(population ~ year, data = .) %>%
       predict(., newdata = tibble(year = c(2020, 2021))) %>%
       tibble(year = c(2020, 2021), 
              population_ext_lm = .)
  ) %>% 
  ungroup() %>% 
  mutate(population_ext_lm = round(population_ext_lm)) %>% 
  arrange(year, GMDNR, sex, nationality, age) %>% 
  relocate(year)
```

```{r include=FALSE}
write_rds(population_ext_lm, "data/BfS/population_ext_lm.Rds")
```

```{r eval=FALSE, include=FALSE}
population_ext_poi <- population %>% 
  filter(year != 2020) %>% 
  # filter(GMDNR == 261 | GMDNR == 782) %>%
  group_by(GMDNR, age, sex, nationality) %>% 
  do(glm(population ~ year, data = ., family = "poisson") %>%
       predict(., newdata = tibble(year = c(2020, 2021)), type = "response") %>%
       tibble(year = c(2020, 2021), 
              population_ext_poi = .)
  ) %>% 
  ungroup() %>% 
  mutate(population_ext_poi = round(population_ext_poi)) %>% 
  arrange(year, GMDNR, sex, nationality, age) %>% 
  relocate(year)
```

```{r eval=FALSE, include=FALSE}
write_rds(population_ext_poi, "data/BfS/population_ext_poi.Rds")
```

Results for large municipality are fine:  

```{r echo=FALSE, warning=FALSE}
population %>% 
  filter(GMDNR == 261 & age == "80+") %>% 
  bind_rows(population_ext_lm %>% 
              filter(GMDNR == 261 & age == "80+")) %>% 
  ggplot(aes(x = year)) + 
  geom_point(aes(y = population)) +
  geom_point(aes(y = population_ext_lm), color = "red") +
  facet_wrap(nationality~sex, scales = "free")
```

```{r eval=FALSE, echo=FALSE, warning=FALSE}
population %>% 
  filter(GMDNR == 261 & age == "80+") %>% 
  bind_rows(population_ext_poi %>% 
              filter(GMDNR == 261 & age == "80+")) %>% 
  ggplot(aes(x = year)) + 
  geom_point(aes(y = population)) +
  geom_point(aes(y = population_ext_poi), color = "red") +
  facet_wrap(nationality~sex, scales = "free")
```

For small municipality we might end up with sth more wiggly.  

```{r echo=FALSE, warning=FALSE}
population %>% 
  filter(GMDNR == 782 & age == "80+") %>% 
  bind_rows(population_ext_lm %>% 
              filter(GMDNR == 782 & age == "80+")) %>% 
  ggplot(aes(x = year)) + 
  geom_point(aes(y = population)) +
  geom_point(aes(y = population_ext_lm), color = "red") +
  facet_wrap(nationality~sex, scales = "free")
```

```{r eval=FALSE,echo=FALSE, warning=FALSE}
population %>% 
  filter(GMDNR == 782 & age == "80+") %>% 
  bind_rows(population_ext_poi %>% 
              filter(GMDNR == 782 & age == "80+")) %>% 
  ggplot(aes(x = year)) + 
  geom_point(aes(y = population)) +
  geom_point(aes(y = population_ext_poi), color = "red") +
  facet_wrap(nationality~sex, scales = "free")
```

## Getting mid-year pops

For each age, sex, nationality stratum, for each municipality separately, simple mean of the two years of data is used to estimate the middle point. In such case we use, for instance data from 31st Dec 2014 and 31st Dec 2015 to estimate population mid-2015.   

```{r}
data <- population %>% 
  filter(year < 2020) %>% 
  bind_rows(population_ext_lm) %>% 
  mutate(population = if_else(year >= 2020, population_ext_lm, population)) %>% 
  select(-population_ext_lm)

for (i in 2014:2020){
  

  result <- data %>% 
    filter(year >= i & year <= i + 1) %>% 
    filter(GMDNR == 261 | GMDNR == 782) %>%
    group_by(GMDNR, age, sex, nationality) %>% 
    summarise(population_mid = mean(population)) %>% 
    ungroup() %>% 
    mutate(year = as.integer(i + 1),
           population_mid = round(population_mid)) %>% 
    arrange(year, GMDNR, sex, nationality, age)
  
  if (i == 2014) {
    population_mid <- result
  } else {
    population_mid <- bind_rows(population_mid, result)
  }
  
}
```

```{r include=FALSE}
rm(i, data, result)
write_rds(population_mid, "data/BfS/population_mid.Rds")
```

Again, results for large municipality are fine:  

```{r echo=FALSE, warning=FALSE}
population %>% 
  filter(year < 2020) %>% 
  bind_rows(population_ext_lm) %>% 
  mutate(population = if_else(year >= 2020, population_ext_lm, population)) %>% 
  select(-population_ext_lm) %>% 
  filter(GMDNR == 261 & age == "80+") %>% 
  bind_rows(population_mid %>% 
              filter(GMDNR == 261 & age == "80+") %>% 
              mutate(year = year - .5)) %>% 
  ggplot(aes(x = year)) + 
  geom_point(aes(y = population)) +
  geom_point(aes(y = population_mid), color = "red") +
  facet_wrap(nationality~sex, scales = "free")
```

Again, for small municipality we might end up with sth more wiggly.  

```{r echo=FALSE, warning=FALSE}
population %>% 
  filter(year < 2020) %>% 
  bind_rows(population_ext_lm) %>% 
  mutate(population = if_else(year >= 2020, population_ext_lm, population)) %>% 
  select(-population_ext_lm) %>% 
  filter(GMDNR == 782 & age == "80+") %>% 
  bind_rows(population_mid %>% 
              filter(GMDNR == 782 & age == "80+") %>% 
              mutate(year = year - .5)) %>% 
  ggplot(aes(x = year)) + 
  geom_point(aes(y = population)) +
  geom_point(aes(y = population_mid), color = "red") +
  facet_wrap(nationality~sex, scales = "free")
```

## Deaths <> population link

Prepared population file is merged to deaths prepared in `02.Rmd`. We use file based on place of residence.  

Non-existent population count `communityXyearXageXsexXnationality` strata are assumed to mean 0 pops.  

```{r}
w_deaths_2015_2021_pop <- read_rds("data/BfS-closed/w_deaths_2015_2021_exp.Rds") %>% 
  left_join(
    bind_rows(population, 
              population %>% 
                filter(year == 2020) %>% 
                mutate(year = 2021))
  ) %>% 
  # left_join(population_ext_poi) %>% 
  # left_join(population_ext_no20) %>%
  replace_na(list(population = 0)) 
```

## Deaths > population problem

```{r}
w_deaths_2015_2021_surplus <- w_deaths_2015_2021_pop %>% 
  filter(deaths > population) %>% 
  arrange(GMDNR, date) %>% 
  mutate(difference = population - deaths)
```

There are `r number(nrow(w_deaths_2015_2021_surplus))` strata from `r number(length(unique(w_deaths_2015_2021_surplus$GMDNR)))`  communities where count of deaths is larger than count of population. The difference is in range of 1-2 and refers to very small counts of deaths/pops.   

```{r echo=FALSE}
w_deaths_2015_2021_surplus %>% 
  select(-GMDNR, -date, -border, -starts_with("AR")) %>% 
  datatable()
```

```{r eval=FALSE, include=FALSE}
# missing checks
w_deaths_2015_2021_exp <- read_rds("data/BfS-closed/w_deaths_2015_2021_exp.Rds")

missing1 <- w_deaths_2015_2021_exp %>% 
  anti_join(population) %>% 
  select(GMDNAME) %>% 
  distinct()

missing2 <- population %>% 
  filter(year >2014) %>% 
  anti_join(w_deaths_2015_2021_exp) %>% 
  select(GMDNAME) %>% 
  distinct()

rm(w_deaths_2015_2021_exp, missing1, missing2); gc()
```

```{r include=FALSE}
rm(w_deaths_2015_2021_surplus); gc()
```

## Labour market aggregates

Prepared population was also aggregated to labour market regions and joined to deaths.    

```{r}
raum <- read_rds("data/BfS/AR.Rds") %>% 
  select(GMDNR, ARNR, ARNAME)

population_ar <- population %>% 
  left_join(raum) %>% 
  group_by(year, ARNR, ARNAME, sex, nationality, age) %>% 
  summarise(population = sum(population)) %>% 
  ungroup()

w_deaths_2015_2021_ar_pop <- read_rds("data/BfS-closed/w_deaths_2015_2021_ar.Rds") %>% 
  left_join(population_ar) %>% 
  replace_na(list(population = 0)) 
```

<!-- ----------------------------------------------------- -->

# Cartogram

2020 population counts (*ignoring age structure*) were used to derive cartograms.  

## Communities

```{r include=FALSE}
# prepared in 01.Rmd
gg21 <- read_rds("data/BfS/gg21.Rds") %>% select(GMDNR)
se21 <- read_rds("data/BfS/se21.Rds") %>% select(GMDNR, GMDNAME)
kt21 <- read_rds("data/BfS/kt21.Rds") %>% select(KTNR, KTNAME)

st_gg21 <- read_rds("data/swisstopo/st_gg21.Rds") %>% 
  st_transform(crs = 2056) %>% 
  select(-GEM_TEIL)
```

```{r eval=FALSE}
carto_data <- 
  gg21 %>% 
  left_join(
    population %>% 
      filter(year == 2020) %>% 
      group_by(GMDNR) %>% 
      summarise(population = sum(population)) %>% 
      ungroup()
  ) 

carto_data %>% 
  select(-BZNR, -KTNR) %>% 
  st_write("data/carto/in_gg21.shp", delete_dsn = TRUE)

se21 %>% 
  select(GMDNR, GMDNAME) %>% 
  st_write("data/carto/in_se21.shp", delete_dsn = TRUE)

# crashing :/
# carto <- carto_data %>%
#   cartogram::cartogram_cont("population")

# write_rds(carto, "data/carto/carto.Rds")

carto_ncont <- carto_data %>% 
  cartogram::cartogram_ncont("population")

write_rds(carto_ncont, "data/carto/carto_ncont.Rds")

carto_dorling <- carto_data %>% 
  cartogram::cartogram_dorling("population")

write_rds(carto_dorling, "data/carto/carto_dorling.Rds")
```

```{r eval=FALSE, include=FALSE}
# checking on swisstopo
# carto_data <- 
#   st_gg21 %>% 
#   select(GMDNR) %>% 
#   left_join(
#     population %>% 
#       filter(year == 2020) %>% 
#       group_by(GMDNR) %>% 
#       summarise(population = sum(population)) %>% 
#       ungroup()
#   )

# crashing :/
# carto <- carto_data %>%
#   cartogram::cartogram_cont("population")
```

```{r eval=FALSE}
# ScapeToad version
# java -Xmx1g -jar "C:\Program Files\ScapeToad-v11\ScapeToad.jar"
```

```{r include=FALSE}
# carto <- read_rds("data/carto/carto.Rds")
carto_ncont <- read_rds("data/carto/carto_ncont.Rds")
carto_dorling <- read_rds("data/carto/carto_dorling.Rds")

# ScapeToad version
carto <- st_read("data/carto/out_gg21.shp")
carto_se <- st_read("data/carto/out_se21.shp")

tmap_mode("plot")
```

### 'Classic' cartogram  

```{r echo=FALSE, message=FALSE, warning=FALSE}
tm_shape(carto) + 
  tm_polygons("population", 
              style = "bclust", n = 5,
              palette = "-RdYlGn",
              legend.show = TRUE) +
  tm_shape(carto_se) + 
  tm_polygons("dodgerblue", 
              legend.show = TRUE, border.col = NA, alpha = 0.5) +
  tm_layout(frame = FALSE)
```

### Noncontinuous cartogram  

```{r echo=FALSE, message=FALSE, warning=FALSE}
tm_shape(kt21) + 
  tm_borders() + 
  tm_shape(carto_ncont) + 
  tm_fill("population", 
          style = "bclust", n = 5,
          palette = "-RdYlGn",
          legend.show = TRUE) +
  tm_layout(frame = FALSE)
```

### Dorling carotgram  

```{r echo=FALSE, message=FALSE, warning=FALSE}
tm_shape(carto_dorling) + 
  tm_fill("population", 
          style = "bclust", n = 5,
          palette = "-RdYlGn",
          legend.show = TRUE) +
  tm_layout(frame = FALSE)
```

```{r include=FALSE}
rm(carto, carto_dorling, carto_ncont, carto_se); gc()
```

## Labour market regions

```{r include=FALSE}
# prepared in 01.Rmd
ar_21_cor <- st_read("data/geo/ar_21_cor.shp") %>% select(-ARNAME)

st_ar_21_cor <- st_read("data/geo/st_ar_21_cor.shp") %>% select(-ARNAME)
```

```{r eval=FALSE}
ar_carto_data <- 
  ar_21_cor %>% 
  left_join(
    population %>% 
      filter(year == 2020) %>% 
      group_by(GMDNR) %>% 
      summarise(population = sum(population)) %>% 
      ungroup() %>% 
      left_join(raum) %>% 
      group_by(ARNR, ARNAME) %>% 
      summarise(population = sum(population)) %>% 
      ungroup()     
  ) 

ar_carto <- ar_carto_data %>%
  cartogram::cartogram_cont("population", itermax = 50)

ar_carto_ncont <- ar_carto_data %>% 
  cartogram::cartogram_ncont("population")

ar_carto_dorling <- ar_carto_data %>% 
  cartogram::cartogram_dorling("population")

write_rds(ar_carto, "data/carto/ar_carto.Rds")
write_rds(ar_carto_ncont, "data/carto/ar_carto_ncont.Rds")
write_rds(ar_carto_dorling, "data/carto/ar_carto_dorling.Rds")
```

```{r include=FALSE}
ar_carto <- read_rds("data/carto/ar_carto.Rds")
ar_carto_ncont <- read_rds("data/carto/ar_carto_ncont.Rds")
ar_carto_dorling <- read_rds("data/carto/ar_carto_dorling.Rds")
```

### 'Classic' cartogram  

```{r echo=FALSE, message=FALSE, warning=FALSE}
tm_shape(ar_carto) + 
  tm_polygons("population", 
              style = "bclust", n = 5,
              palette = "-RdYlGn",
              legend.show = TRUE) +
  tm_layout(frame = FALSE)
```

### Noncontinuous cartogram  

```{r echo=FALSE, message=FALSE, warning=FALSE}
tm_shape(kt21) + 
  tm_borders() + 
  tm_shape(ar_carto_ncont) + 
  tm_fill("population", 
          style = "bclust", n = 5,
          palette = "-RdYlGn",
          legend.show = TRUE) +
  tm_layout(frame = FALSE)
```

### Dorling carotgram  

```{r echo=FALSE, message=FALSE, warning=FALSE}
tm_shape(ar_carto_dorling) + 
  tm_fill("population", 
          style = "bclust", n = 5,
          palette = "-RdYlGn",
          legend.show = TRUE) +
  tm_layout(frame = FALSE)
```

```{r include=FALSE}
rm(ar_carto, ar_carto_ncont, ar_carto_dorling); gc()
```

<!-- ----------------------------------------------------- -->

# Swiss-SEP

```{r}
sep3 <- read_rds("../SNC_Swiss-SEP2/FINAL/RDS/ssep3_user_geo.Rds") %>%
  select(gisid, ssep3, ssep3_d) %>% 
  st_transform(crs = 2056)
```

```{r eval=FALSE, include=FALSE}
# %>%
#   mutate(ssep3_d = factor(ssep3_d,
#                           levels = 1:10,
#                           labels = c("1st - lowest",
#                                      "2", "3", "4",
#                                      "5th decile",
#                                      "6", "7", "8", "9",
#                                      "10th - highest")))
```

Using `r number(nrow(sep3), big.mark = ",")` n'hoods from version 3.0 (hence slightly unequal size of deciles!):  

```{r echo=FALSE}
sep3 %>% 
  st_drop_geometry() %>% 
  frq(ssep3_d)
```

This is aggregated to the level of `r scales::number(length(unique(st_gg21$GMDNR)), big.mark = ",")` communities from swisstopo.

## STATPOP

```{r}
r19_bu_orig <- read_dta("../SNC_core/data-raw/statpop/r19_bu_orig.dta") %>% 
  zap_label() %>% zap_formats()

r19_bu <- r19_bu_orig %>% 
  select(r19_egid_snc, r19_comm, r19_gdename,
         r19_ecoord, r19_ncoord,
         r19_nrpers_main) %>% 
  # select(r19_nrpers_total, r19_nrpers_perm) %>% 
  # filter(!is.na(r19_ncoord)) %>% 
  # filter(!is.na(r19_ecoord)) %>% 
  filter(r19_nrpers_main > 0) %>% 
  st_as_sf(coords = c("r19_ecoord", "r19_ncoord"), 
           crs = 2056,
           remove = TRUE)
```

Using data from `STATPOP` **2019**.  

Used dataset consists of `r scales::number(nrow(r19_bu), big.mark = ",")` buildings with at least one person registered inside. Variable `r19_nrpers_main` was used for this estimate.^[Slightly different populations would be obtained with different conceptual models of *residential populations* by using either `r19_nrpers_total` or `r19_nrpers_perm` variables!] Population of CH estimated from this dataset is `r scales::number(sum(r19_bu$r19_nrpers_main), big.mark = ",")`.   

The nearest building with SSEP was assigned for each of these building. The point dataset was then overlaid with community boundaries and summary measures of population based index were created and population level SEP was aggregated by community. 

```{r include=FALSE}
rm(r19_bu_orig); gc()
```

## 'nhood based averages for communities

```{r eval=FALSE}
st_gg21 <- read_rds("data/swisstopo/st_gg21.Rds") %>% 
  st_transform(crs = 2056) %>% 
  select(GMDNR)

ssep3_gem_21 <- 
  st_join(sep3, 
          st_gg21, 
          join = st_intersects) 

# tm_shape(gg21) +
#   tm_polygons() +
#   tm_shape(ssep3_gem_21 %>% filter(is.na(GMDNR))) +
#   tm_dots()

ssep3_gem_21_neigh <- ssep3_gem_21 %>% 
  st_drop_geometry() %>% 
  group_by(GMDNR) %>% 
  summarise(n = n(),
            # ssep3_mean = mean(ssep3),
            ssep3_median = median(ssep3)) %>% 
  ungroup() 

ssep3_gem_21_neigh_geo <- inner_join(gg21, ssep3_gem_21_neigh) %>% 
  filter(!is.na(ssep3_median)) %>% 
  mutate(median_ssep3_d = ntile(ssep3_median, 10)) %>% 
  mutate(median_ssep3_d = factor(median_ssep3_d,
                                 levels = 1:10,
                                 labels = c("1st - lowest", 
                                            "2", "3", "4", 
                                            "5th decile", 
                                            "6", "7", "8", "9", 
                                            "10th - highest")))

write_rds(ssep3_gem_21_neigh_geo, "data/sep/ssep3_gem_21_neigh_geo.rds")
```

```{r include=FALSE}
ssep3_gem_21_neigh_geo <- read_rds("data/sep/ssep3_gem_21_neigh_geo.rds")
```

Watch out for small n! Here some examples with communities below 20 n'hoods

```{r echo=FALSE}
ssep3_gem_21_neigh_geo %>% 
  st_drop_geometry() %>% 
  select(GMDNR, GMDNAME, n) %>% 
  inner_join(st_drop_geometry(st_gg21)) %>% 
  select(-GMDNR) %>% 
  filter(n < 20) %>%
  arrange(n) %>% 
  datatable(rownames = FALSE, options = list(dom = 't'))

# ssep3_gem_21_neigh_geo %>% 
#   filter(n < 20) %>%
#   qtm(borders = "red", fill = "red")
```

Final map using median index - bubbles scaled to number of 'nhoods.

```{r include=FALSE}
tmap_mode("view")
```

```{r eval=FALSE, include=FALSE}
frq(ssep3_gem_21_neigh_geo$median_ssep3_d)
```

```{r echo=FALSE}
# tm_shape(kt21) +
#   tm_borders() + 
tm_shape(ssep3_gem_21_neigh_geo) +
  tm_dots(size = "n", col = "median_ssep3_d", palette = "RdYlGn",
          scale = .5,  shape = 19)
```

## Population based averages for communities

```{r eval=FALSE}
# join to sep3
r19_bu_sep <- st_join(r19_bu, sep3, join = st_nearest_feature)

# fails for unknown and unsolved reasons
# nearest <- st_nearest_feature(r19_bu, sep3)
# r19_bu_sep$dist3 <- st_distance(r19_bu_sep, sep3[nearest, ], by_element = TRUE)
# rm(nearest, r19_bu); gc()
# 
# summary(r19_bu_sep$dist3)

r19_bu_sep_gem21 <- 
  st_join(r19_bu_sep, st_gg21, join = st_intersects) %>%  
  relocate(r19_egid_snc, gisid, r19_nrpers_main,
           ssep3, ssep3_d,
           GMDNR, r19_comm, r19_gdename) 

# View(r19_bu_sep_gem21)

write_rds(r19_bu_sep_gem21, "data/sep/r19_bu_sep_gem21.rds")
```

```{r include=FALSE}
rm(r19_bu); gc()

r19_bu_sep_gem21 <- read_rds("data/sep/r19_bu_sep_gem21.rds") %>% 
  st_drop_geometry()
```

```{r}
ssep3_gem_21_pop <- r19_bu_sep_gem21 %>% 
  select(GMDNR, ssep3, r19_nrpers_main) %>% 
  uncount(r19_nrpers_main) %>% 
  group_by(GMDNR) %>% 
  summarise(pop = n(),
            # ssep3_mean = mean(ssep3),
            ssep3_median = median(ssep3)) %>% 
  ungroup() 

# left_join(gg21, ssep3_gem_21_pop) %>% 
#   filter(is.na(ssep3_median))

ssep3_gem_21_pop_geo <- left_join(st_gg21, ssep3_gem_21_pop) %>% 
  filter(!is.na(ssep3_median)) %>% 
  mutate(median_ssep3_d = ntile(ssep3_median, 10)) %>% 
  mutate(median_ssep3_d = factor(median_ssep3_d,
                                 levels = 1:10,
                                 labels = c("1st - lowest", 
                                            "2", "3", "4", 
                                            "5th decile", 
                                            "6", "7", "8", "9", 
                                            "10th - highest")))

write_rds(ssep3_gem_21_pop_geo, "data/sep/ssep3_gem_21_pop_geo.rds")
```

Final map using median index - bubbles scaled to number of people (2019)  

```{r eval=FALSE, include=FALSE}
frq(ssep3_gem_21_pop_geo$median_ssep3_d)
```

```{r echo=FALSE}
ssep3_gem_21_pop_geo %>% 
  tm_shape() +
  tm_dots(size = "pop", col = "median_ssep3_d", palette = "RdYlGn",
          scale = .5,  shape = 19)
```

## Population based averages for labour market regions

```{r include=FALSE}
# using swisstopo due to bfs geodata errors :/
ar_21 <- st_read("data/geo/st_ar_21_cor.shp") %>% 
  select(-ARNAME)
```

```{r}
ssep3_ar_21_pop <- r19_bu_sep_gem21 %>% 
  select(GMDNR, ssep3, r19_nrpers_main) %>% 
  left_join(raum) %>% 
  select(-GMDNR) %>% 
  uncount(r19_nrpers_main) %>% 
  group_by(ARNR, ARNAME) %>% 
  summarise(pop = n(),
            # ssep3_mean = mean(ssep3),
            ssep3_median = median(ssep3)) %>% 
  ungroup() 

# left_join(ar_21, ssep3_ar_21_pop) %>%
#   filter(is.na(ssep3_median))

ssep3_ar_21_pop_geo <- left_join(ar_21, ssep3_ar_21_pop) %>% 
  filter(!is.na(ssep3_median)) %>% 
  mutate(median_ssep3_d = ntile(ssep3_median, 10)) %>% 
  mutate(median_ssep3_d = factor(median_ssep3_d,
                                 levels = 1:10,
                                 labels = c("1st - lowest", 
                                            "2", "3", "4", 
                                            "5th decile", 
                                            "6", "7", "8", "9", 
                                            "10th - highest")))

write_rds(ssep3_ar_21_pop_geo, "data/sep/ssep3_ar_21_pop_geo.rds")
```

```{r include=FALSE}
rm(r19_bu_sep_gem21); gc()
```

Final map using median index - bubbles scaled to number of people (2019)  

```{r eval=FALSE, include=FALSE}
frq(ssep3_ar_21_pop_geo$median_ssep3_d)
```

```{r echo=FALSE}
ssep3_ar_21_pop_geo %>% 
  tm_shape() +
  tm_dots(size = "pop", col = "median_ssep3_d", palette = "RdYlGn",
          scale = .5,  shape = 19)
```

<!-- ----------------------------------------------------- -->

# Raumgliederungen 

Data from the [app](https://www.agvchapp.bfs.admin.ch/de/typologies/query). State as of `2021-07-01` to match spatial data nicely.  

```{r}
raum <- read_xlsx("data-raw/BfS/Raumgliederungen.xlsx", 
                  skip = 1) %>% 
  remove_empty(c("rows", "cols")) %>% clean_names() %>% 
  filter(! is.na(bfs_gde_nummer)) %>% 
  rename(GMDNR = bfs_gde_nummer,
         GMDNAME = gemeindename,
         KTNAME = kanton,
         BZNR = bezirks_nummer,
         BZNAME = bezirksname) %>% 
  select(GMDNR, GMDNAME, KTNAME, BZNR, BZNAME, 
         stadtische_landliche_gebiete, sprachgebiete,
         urbanisierungsgrad_2011_degurba_eurostat) %>% 
  mutate(
    r_urban1 = case_when(
      stadtische_landliche_gebiete == 1 ~  "Urban",
      stadtische_landliche_gebiete == 2 ~  "Periurban",
      stadtische_landliche_gebiete == 3 ~  "Rural",
      TRUE ~  ""),
    r_urban2 = case_when(
      urbanisierungsgrad_2011_degurba_eurostat == 1 ~  "Dense",
      urbanisierungsgrad_2011_degurba_eurostat == 2 ~  "Medium",
      urbanisierungsgrad_2011_degurba_eurostat == 3 ~  "Low",
      TRUE ~  ""),
    r_lang = case_when(
      sprachgebiete == 1 ~  "German",
      sprachgebiete == 2 ~  "French",
      sprachgebiete == 3 ~  "Italian",
      sprachgebiete == 4 ~  "Romansh",
      TRUE ~  "")
  ) %>% 
  select(-stadtische_landliche_gebiete, -sprachgebiete,
         -urbanisierungsgrad_2011_degurba_eurostat)
```

```{r include=FALSE}
stopifnot(nrow(anti_join(raum, gg21)) == 0)

gg21_merged <- gg21 %>% 
  left_join(raum) %>% 
  relocate(geometry, .after = last_col())

tmap_mode("plot")
```

## Urbanization 1 

```{r echo=FALSE}
frq(raum, r_urban1)

tm_shape(se21, is.master = FALSE) +
  tm_fill(col = "#a7cdf2")  +
  tm_shape(gg21_merged) +
  tm_fill(col = "r_urban1")  +
  tm_shape(kt21) +
  tm_borders()  +
  tm_layout(
    title = "Urbanization 2021"
  )
```

## Urbanization 2 

Using [DEGURBA](https://ec.europa.eu/eurostat/web/degree-of-urbanisation/background).  

```{r echo=FALSE}
frq(raum, r_urban2)

tm_shape(se21, is.master = FALSE) +
  tm_fill(col = "#a7cdf2")  +
  tm_shape(gg21_merged) +
  tm_fill(col = "r_urban2", palette = "-Set3")  +
  tm_shape(kt21) +
  tm_borders()  +
  tm_layout(
    title = "Urbanization 2021"
  )
```

## Language region

```{r echo=FALSE}
frq(raum, r_lang)

tm_shape(se21, is.master = FALSE) +
  tm_fill(col = "#a7cdf2")  +
  tm_shape(gg21_merged) +
  tm_fill(col = "r_lang")  +
  tm_shape(kt21) +
  tm_borders()  +
  tm_layout(
    title = "Language region 2021"
  )
```

```{r include=FALSE}
# final link
w_deaths_2015_2021_pop %<>% 
  left_join(ssep3_gem_21_pop_geo %>% 
              st_drop_geometry() %>% 
              select(-pop)) %>% 
  left_join(gg21_merged %>% 
              st_drop_geometry() %>% 
              select(-GMDNAME, -KTNAME, -BZNR, -BZNAME)
  )

write_rds(w_deaths_2015_2021_pop, "data/BfS-closed/w_deaths_2015_2021_pop.Rds")

w_deaths_2015_2021_ar_pop %>% 
  left_join(
    read_rds("data/BfS/raum.Rds") %>% 
      select(ARNR, border)
  ) %>% 
  left_join(ssep3_ar_21_pop_geo %>% 
              st_drop_geometry() %>% 
              select(-pop)
  )

write_rds(w_deaths_2015_2021_ar_pop, "data/BfS-closed/w_deaths_2015_2021_ar_pop.Rds")
```

