---
title: "Spatial analyses of 2020-21 excess mortality in CH"
subtitle: "Ancillary data preparation"
pagetitle: "Spatial Excess: more data"
author: "Radek Panczak"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
    code_folding: show
    self_contained: true
    highlight: pygments
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../docs") })
---

<!-- ----------------------------------------------------- -->

```{r setup, include = FALSE}
set.seed(12345)
options(scipen = 999)
options(max.print = "75")

library(pacman)
p_load(
  haven, readxl,
  tidyverse, magrittr, janitor, scales, lubridate,
  DT,
  sf, tmap,
  swissdd
)

import::from("sjmisc", "frq")
```

```{r conflicts, include = FALSE}
tidyverse::tidyverse_conflicts()

conflicted::conflict_scout()
```

```{r knit-setup, include = FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

knitr::opts_chunk$set(
  cache = FALSE,
  prompt = FALSE,
  tidy = FALSE,
  comment = NA,
  message = FALSE,
  warning = FALSE
)

knitr::opts_knit$set(width = 75)
```

<!-- ----------------------------------------------------- -->

# Population

Data extracted from [`px-x-0103010000_201`](https://www.pxweb.bfs.admin.ch/pxweb/de/px-x-0103010000_201/-/px-x-0103010000_201.px/):

> Ständige und nichtständige Wohnbevölkerung nach institutionellen Gliederungen, Staatsangehörigkeit (Kategorie), Geburtsort, Geschlecht und Altersklasse

Important info from footnotes:

> *Letzte Änderungen: Neuer Datensatz (Jahr 2020)*
> *Stand der Datenbank: Juni 2021*
> *Stichtag: 31. Dezember*
> **Raumbezug: Gemeinden / 18.10.2020**
> *Datenquelle: Statistik der Bevölkerung und der Haushalte STATPOP*
> *Definition der ständigen Wohnbevölkerung*

## Community codes

Community mutations data after `2021-01-01` from @Cordula are integrated into the data and pops are recalculated using these new codes to match mortality.  

```{r eval=FALSE, include=FALSE}
histcomm_14 <- read_excel("data-raw/BfS/Gemeindestand - Stand vom 01.07.2021.xlsx",
                          col_types = c(
                            "numeric", "text", "skip",
                            "skip", "numeric", "text", "text"
                          )
) %>%
  remove_empty(c("rows", "cols")) %>%
  clean_names() %>%
  mutate(datum_der_aufnahme = ymd(datum_der_aufnahme)) %>%
  filter(datum_der_aufnahme >= ymd("2014-01-01") &
           datum_der_aufnahme < ymd("2015-01-01")) %>%
  arrange(desc(datum_der_aufnahme), gemeindename)
```

```{r}
histcomm <- read_rds("data/BfS/histcomm.Rds") %>%
  filter(datum_der_aufnahme >= ymd("2021-01-01")) %>%
  select(bfs_gde_num_old, bfs_gde_num_new, gemeindename_new)
```

```{r eval=FALSE, include=FALSE}
datatable(histcomm)
```

## Data preps

Data aggregated to age groups as in deaths dataset.  

```{r}
population <- read_xlsx("data-raw/BfS/px-x-0103010000_201.xlsx",
                        col_types = c(
                          "numeric",
                          "skip", "numeric", "text", "skip",
                          "skip", "skip", "text", "skip", "skip",
                          "skip", "text", "numeric", "numeric",
                          "numeric", "numeric", "numeric",
                          "numeric", "numeric", "numeric",
                          "numeric", "numeric", "numeric",
                          "numeric", "numeric", "numeric",
                          "numeric", "numeric", "numeric",
                          "numeric", "numeric", "numeric",
                          "numeric"
                        ),
                        skip = 1
) %>%
  remove_empty(c("rows", "cols")) %>%
  clean_names() %>%
  rename(
    year = x1,
    GMDNR = x2,
    GMDNAME = x3
  ) %>%
  mutate(
    sex = if_else(x5 == "Frau", "Female", "Male"),
    nationality = if_else(x4 == "Schweiz", "Swiss", "Foreigner"),
    year = as.integer(year)
  ) %>%
  select(-x4, -x5) %>%
  relocate(sex, .after = GMDNAME) %>%
  relocate(nationality, .after = sex) %>%
  mutate(GMDNAME = word(GMDNAME, 2, -1)) %>%
  fill(year, GMDNR, GMDNAME, nationality) %>%
  mutate(`<40` = x0_4_jahre + x5_9_jahre + x10_14_jahre + x15_19_jahre +
           x20_24_jahre + x25_29_jahre + x30_34_jahre + x35_39_jahre) %>%
  select(
    -x0_4_jahre, -x5_9_jahre, -x10_14_jahre, -x15_19_jahre,
    -x20_24_jahre, -x25_29_jahre, -x30_34_jahre, -x35_39_jahre
  ) %>%
  mutate(`40-49` = x40_44_jahre + x45_49_jahre) %>%
  mutate(`50-59` = x50_54_jahre + x55_59_jahre) %>%
  mutate(`60-69` = x60_64_jahre + x65_69_jahre) %>%
  mutate(`70-79` = x70_74_jahre + x75_79_jahre) %>%
  select(
    -x40_44_jahre, -x45_49_jahre, -x50_54_jahre, -x55_59_jahre,
    -x60_64_jahre, -x65_69_jahre, -x70_74_jahre, -x75_79_jahre
  ) %>%
  mutate(`80+` = x80_84_jahre + x85_89_jahre + x90_94_jahre +
           x95_99_jahre + x100_jahre_und_mehr) %>%
  select(
    -x80_84_jahre, -x85_89_jahre, -x90_94_jahre,
    -x95_99_jahre, -x100_jahre_und_mehr
  ) %>%
  pivot_longer(
    cols = `<40`:`80+`,
    names_to = "age", values_to = "population"
  ) %>%
  left_join(histcomm, by = c("GMDNR" = "bfs_gde_num_old")) %>%
  mutate(
    GMDNR = if_else(!is.na(bfs_gde_num_new), bfs_gde_num_new, GMDNR),
    GMDNAME = if_else(!is.na(gemeindename_new), gemeindename_new, GMDNAME)
  ) %>%
  select(-bfs_gde_num_new, -gemeindename_new) %>%
  group_by(year, GMDNR, GMDNAME, sex, age) %>%
  summarise(population = as.integer(sum(population))) %>%
  ungroup() %>%
  arrange(year, GMDNR, sex, age)
```

```{r include=FALSE}
rm(histcomm); gc()
```

Yearly totals (**on 31st Dec**!):  

```{r echo=FALSE}
population %>%
  group_by(year) %>%
  summarise(population = sum(population)) %>%
  mutate(population = as.character(number(population, big.mark = ",")))
```

Age (40 plus!) distributions over years  

```{r echo=FALSE}
population %>%
  filter(age != "<40") %>%
  group_by(year, age, sex) %>%
  summarise(population = sum(population)) %>%
  ungroup() %>%
  ggplot(aes(x = age, y = population, fill = as.factor(year))) +
  geom_col(position = "dodge") +
  scale_fill_viridis_d()
```

## Predicting 2020 & 2021 pops

For each age, sex, nationality stratum, for each municipality separately, Poisson model was fitted using the 2014-2019 data; then prediction was made for 2020 & 2021 years. The gist of this operation was to exclude the effect of pandemic from the population counts in years affected. 

```{r eval=FALSE, include=FALSE}
# smallest pops
population %>%
  filter(year == 2020) %>%
  group_by(GMDNR) %>%
  summarize(
    population = sum(population),
    GMDNAME = first(GMDNAME)
  ) %>%
  ungroup() %>%
  arrange(population)
# arrange(desc(population))
```

```{r}
population_ext_poi <- population %>%
  filter(year != 2020) %>%
  # filter(GMDNR == 261 | GMDNR == 389) %>%
  group_by(GMDNR, age, sex) %>%
  do(glm(population ~ year, data = ., family = "poisson") %>%
       predict(., newdata = tibble(year = c(as.integer(2020), as.integer(2021))), type = "response") %>%
       tibble(
         year = c(as.integer(2020), as.integer(2021)),
         population_ext_poi = .
       )) %>%
  ungroup() %>%
  mutate(population_ext_poi = as.integer(round(population_ext_poi))) %>%
  arrange(year, GMDNR, sex, age) %>%
  relocate(year)
```

Summary of values:  

```{r echo=FALSE}
summary(population_ext_poi$population_ext_poi)
```

**Note:** *4 NAs!*  

```{r echo=FALSE}
population %>%
  filter(GMDNR %in% c(4232) & sex == "Male" & age == "80+")
population_ext_poi %>%
  filter(GMDNR %in% c(4232) & sex == "Male" & age == "80+")

population %>%
  filter(GMDNR %in% c(389) & sex == "Male" & age == "80+")
population_ext_poi %>%
  filter(GMDNR %in% c(389) & sex == "Male" & age == "80+")
```

### Results in large municipality

Results for large municipality are fine.    

```{r echo=FALSE, warning=FALSE}
population %>%
  filter(GMDNR == 261 & (age == "80+" | age == "<40")) %>%
  bind_rows(population_ext_poi %>%
              filter(GMDNR == 261 & (age == "80+" | age == "<40"))) %>%
  ggplot(aes(x = year)) +
  geom_point(aes(y = population)) +
  geom_point(aes(y = population_ext_poi), color = "red") +
  facet_wrap(age ~ sex, scales = "free")
```

### Results in small municipality

For small municipality we might end up with sth more wiggly. 

```{r echo=FALSE, warning=FALSE}
population %>%
  filter(GMDNR == 389 & (age == "80+" | age == "<40")) %>%
  bind_rows(population_ext_poi %>%
              filter(GMDNR == 389 & (age == "80+" | age == "<40"))) %>%
  ggplot(aes(x = year)) +
  geom_point(aes(y = population)) +
  geom_point(aes(y = population_ext_poi), color = "red") +
  facet_wrap(age ~ sex, scales = "free")
```

**Note lack of prediction for 2020 & 2021 for oldest males!**  

There are two municipalities with missing predictions for oldest males:  

```{r echo=FALSE}
(problems <- population_ext_poi %>%
   filter(is.na(population_ext_poi)) %>%
   select(GMDNR, age, sex) %>%
   left_join(population %>%
               select(GMDNR, GMDNAME) %>%
               distinct()) %>%
   distinct())
```

```{r eval=FALSE, include=FALSE}
population %>%
  bind_rows(population_ext_poi) %>%
  filter(
    GMDNR %in% problems$GMDNR,
    age %in% problems$age,
    sex %in% problems$sex
  ) %>%
  arrange(GMDNR, sex, age, year) %>%
  View()

population %>%
  bind_rows(population_ext_lm) %>%
  filter(
    GMDNR %in% problems$GMDNR,
    age %in% problems$age,
    sex %in% problems$sex
  ) %>%
  arrange(GMDNR, sex, age, year) %>%
  View()
```

Population in this strata was replaced by estimates of a simple linear model.  

```{r}
population_ext_lm <- population %>%
  filter(year != 2020) %>%
  # filter(GMDNR == 261 | GMDNR == 389) %>%
  group_by(GMDNR, age, sex) %>%
  do(lm(population ~ year, data = .) %>%
       predict(., newdata = tibble(year = c(as.integer(2020), as.integer(2021)))) %>%
       tibble(
         year = c(as.integer(2020), as.integer(2021)),
         population_ext_lm = .
       )) %>%
  ungroup() %>%
  mutate(population_ext_lm = as.integer(round(population_ext_lm))) %>%
  arrange(year, GMDNR, sex, age) %>%
  relocate(year)
```

```{r eval=FALSE, include=FALSE}
population %>%
  bind_rows(population_ext_lm) %>%
  filter(
    GMDNR %in% problems$GMDNR,
    age %in% problems$age,
    sex %in% problems$sex
  ) %>%
  arrange(GMDNR, sex, age, year) %>%
  View()
```

## Getting mid-year pops 

For each age, sex, nationality stratum, for each municipality separately, simple mean of the two years of data is used to estimate the middle point. In such case we use, for instance data from 31st Dec 2014 and 31st Dec 2015 to estimate population mid-2015.   

```{r eval=FALSE, include=FALSE}
table(population$year)
table(population_ext_poi$year)
```

```{r}
data <- population %>%
  filter(year < 2020) %>%
  bind_rows(population_ext_poi) %>%
  # 2020 replaced by estimates from poisson
  mutate(population = if_else(year >= 2020,
                              population_ext_poi, population
  )) %>%
  # replacing two missing strata with lm estimates
  left_join(population_ext_lm) %>%
  mutate(population = if_else(year >= 2020 & is.na(population_ext_poi),
                              population_ext_lm, population
  )) %>%
  select(-population_ext_poi, -population_ext_lm)

for (i in 2014:2020) {
  result <- data %>%
    filter(year >= i & year <= i + 1) %>%
    # filter(GMDNR == 261 | GMDNR == 389) %>%
    group_by(GMDNR, age, sex) %>%
    summarise(population_mid_poi = mean(population)) %>%
    ungroup() %>%
    mutate(
      year = as.integer(i + 1),
      population_mid_poi = as.integer(round(population_mid_poi))
    ) %>%
    arrange(year, GMDNR, sex, age)
  
  if (i == 2014) {
    population_mid_poi <- result
  } else {
    population_mid_poi <- bind_rows(population_mid_poi, result)
  }
}
```

```{r include=FALSE}
table(data$year)
table(population_mid_poi$year)

rm(i, data, result, population_ext_lm)

population_mid_poi %<>%
  arrange(year, GMDNR, sex, age) %>%
  relocate(year)
```

### Large municipality 

Again, results for large municipality are fine:  

```{r echo=FALSE, warning=FALSE}
population %>%
  filter(year < 2020) %>%
  bind_rows(population_ext_poi) %>%
  mutate(population = if_else(year >= 2020, population_ext_poi, population)) %>%
  select(-population_ext_poi) %>%
  filter(GMDNR == 261 & age == "80+") %>%
  bind_rows(population_mid_poi %>%
              filter(GMDNR == 261 & age == "80+") %>%
              mutate(year = year - .5)) %>%
  ggplot(aes(x = year)) +
  geom_point(aes(y = population)) +
  geom_point(aes(y = population_mid_poi), color = "red") +
  facet_wrap(~sex, scales = "free")
```

### Small municipality 

Again, for small municipality we might end up with sth more wiggly.  

```{r echo=FALSE, warning=FALSE}
population %>%
  filter(year < 2020) %>%
  bind_rows(population_ext_poi) %>%
  mutate(population = if_else(year >= 2020, population_ext_poi, population)) %>%
  select(-population_ext_poi) %>%
  filter(GMDNR == 389 & age == "80+") %>%
  bind_rows(population_mid_poi %>%
              filter(GMDNR == 389 & age == "80+") %>%
              mutate(year = year - .5)) %>%
  ggplot(aes(x = year)) +
  geom_point(aes(y = population)) +
  geom_point(aes(y = population_mid_poi), color = "red") +
  facet_wrap(~sex, scales = "free")
```

## Deaths <-> population link

Prepared population file is merged to deaths dataset prepared in `02.Rmd`.  

**We use file based on place of residence.**  

Non-existent population count `communityXyearXageXsexXnationality` strata are assumed to mean 0 pops.  

```{r}
w_deaths_2015_2021_pop <- read_rds("data/BfS-closed/monthly_deaths/w_deaths_2015_2021_exp.Rds") %>%
  left_join(population) %>% 
  mutate(population = ifelse(year < 2021 & is.na(population), 0, population)) %>% 
  left_join(population_mid_poi) %>% 
  mutate(population_mid_poi = ifelse(is.na(population_mid_poi), 0, population_mid_poi))
```

```{r eval=FALSE, include=FALSE}
summary(w_deaths_2015_2021_pop$population)
summary(w_deaths_2015_2021_pop$population_mid_poi)
```

## Deaths > population problem

```{r}
surplus <- w_deaths_2015_2021_pop %>%
  filter(deaths > population) %>%
  arrange(GMDNR, date) %>%
  mutate(difference = population - deaths)
```

There are `r number(nrow(surplus))` strata from `r number(length(unique(surplus$GMDNR)))`  communities where count of deaths is larger than count of population. The difference is in range of 1-2 and refers to very small counts of deaths/pops.   

```{r echo=FALSE}
surplus %>%
  select(-GMDNR, -date, -border, -starts_with("AR")) %>%
  datatable()
```

```{r eval=FALSE, include=FALSE}
# missing checks
w_deaths_2015_2021_exp <- read_rds("data/BfS-closed/monthly_deaths/w_deaths_2015_2021_exp.Rds")

missing1 <- w_deaths_2015_2021_exp %>%
  anti_join(population) %>%
  select(GMDNAME) %>%
  distinct()

missing2 <- population %>%
  filter(year > 2014) %>%
  anti_join(w_deaths_2015_2021_exp) %>%
  select(GMDNAME) %>%
  distinct()

rm(w_deaths_2015_2021_exp, missing1, missing2)
gc()
```

```{r include=FALSE}
rm(surplus); gc()

write_rds(w_deaths_2015_2021_pop, "data/BfS-closed/monthly_deaths/w_deaths_2015_2021_pop.Rds")
```

<!-- ----------------------------------------------------- -->

## STATPOP

```{r}
statpop2020 <- read_delim("data-raw/BfS-closed/STATPOP/statpop2020_220098p.zip", 
                          delim = ";", escape_double = FALSE, 
                          col_types = cols(STATYEAR = col_integer(), 
                                           SEX = col_integer(), 
                                           TYPEOFRESIDENCE = col_integer(), 
                                           POPULATIONTYPE = col_integer(), 
                                           AGE = col_integer(), 
                                           CLASSAGEFIVEYEARS = col_integer(), 
                                           NATIONALITYCATEGORY = col_integer(), 
                                           MAINRESIDENCECATEGORY = col_integer(), 
                                           GEOCOORDE = col_integer(), 
                                           GEOCOORDN = col_integer(), 
                                           INDIC_EGID = col_integer(), 
                                           statdate = col_date(format = "%d/%m/%Y")), 
                          trim_ws = TRUE) %>% 
  janitor::remove_empty(c("rows", "cols")) %>% janitor::clean_names()  

statpop2020_agg <- statpop2020 %>% 
  rename(egid = federalbuildingid) %>% 
  filter(typeofresidence == 1) %>% 
  filter(populationtype == 1) %>% 
  filter(mainresidencecategory == 1) %>% 
  filter(indic_egid == 1) %>% 
  select(-statyear, -statdate, 
         -typeofresidence, -populationtype, -mainresidencecategory,
         -indic_egid) %>% 
  group_by(egid, geocoorde, geocoordn) %>% 
  summarise(pop = n()) %>% 
  ungroup() %>% 
  st_as_sf(coords = c("geocoorde", "geocoordn"), 
           crs = 2056,
           remove = TRUE)
```

Using data from `STATPOP` **2020**.  

Dataset consists of `r scales::number(nrow(statpop2020), big.mark = ",")` individuals and `r scales::number(nrow(statpop2020_agg), big.mark = ",")` buildings. We selected individuals with type of residence *Hauptwohnsitz*, population type *Ständige Wohnbevölkerung*, main residency category *Nur ein Hauptwohnsitz* and with valid EGID building ID.  

```{r include=FALSE}
write_rds(statpop2020_agg, "data/BfS-closed/STATPOP/statpop2020_agg.Rds")

rm(statpop2020); gc()
```

<!-- ----------------------------------------------------- -->

# Voting

Data from `swissdd` [package](https://github.com/politanch/swissdd).    

Linked to municipality boundaries from `2022-01-01`.  

```{r}
st_gg22 <- read_rds("data/swisstopo/st_gg22.Rds")
```

## June vote

> Bundesgesetz vom 25.09.2020 über die gesetzlichen Grundlagen für Verordnungen des Bundesrates zur Bewältigung der Covid-19-Epidemie (Covid-19-Gesetz)

### Data 

```{r}
covid_jun <- get_nationalvotes(
  votedates = "2021-06-13",
  geolevel = "municipality"
) %>%
  filter(name == "Bundesgesetz über die gesetzlichen Grundlagen für Verordnungen des Bundesrates zur Bewältigung der Covid-19-Epidemie (Covid-19-Gesetz)") %>%
  select(-name, -id)
```

```{r echo=FALSE, include=FALSE}
# Votes from abroad are excluded.
# covid_jun %>%
#   filter(str_detect(mun_name, fixed("Ausland")) | str_detect(mun_name, fixed("l'étranger"))) %>%
#   select(mun_name, jaStimmenAbsolut, jaStimmenInProzent)

covid_jun %<>%
  filter(!str_detect(mun_name, fixed("Ausland"))) %>%
  filter(!str_detect(mun_name, fixed("l'étranger"))) %>%
  mutate(GMDNR = as.numeric(mun_id)) %>%
  relocate(GMDNR)
```

Dataset consists of `r scales::number(length(unique(covid_jun$mun_id)), big.mark = ",")` communities.

### Results preview

```{r echo=FALSE}
plot_nationalvotes(
  votedate = "2021-06-13",
  vote_id = 6430, geolevel = "municipality"
)
```

### Link to municipalities

Using `GMDNR` / `mun_id` for deterministic linkage.  

```{r}
covid_jun_gem <- st_gg22 %>%
  left_join(covid_jun)
```

There are few communities that exist on the map but do not exist in the voting dataset:   

```{r echo=FALSE}
covid_jun_gem %>%
  st_drop_geometry() %>%
  filter(is.na(jaStimmenInProzent)) %>%
  select(GMDNR, GMDNAME)
```

```{r include=FALSE}
write_rds(covid_jun, "data/voting/covid_jun.Rds")
write_rds(covid_jun_gem, "data/voting/covid_jun_gem.Rds")
```

## November vote

> Änderung des Bundesgesetzes über die gesetzlichen Grundlagen für Verordnungen des Bundesrates zur Bewältigung der Covid-19-Epidemie (Covid-19-Gesetz) (Härtefälle, Arbeitslosenversicherung, familienergänzende Kinderbetreuung, Kulturschaffende, Veranstaltungen)  

### Data 

```{r}
covid_nov <- get_nationalvotes(
  votedates = "2021-11-28",
  geolevel = "municipality"
) %>%
  filter(name == "Änderung des Bundesgesetzes über die gesetzlichen Grundlagen für Verordnungen des Bundesrates zur Bewältigung der Covid-19-Epidemie (Covid-19-Gesetz) (Härtefälle, Arbeitslosenversicherung, familienergänzende Kinderbetreuung, Kulturschaffende, Veranstaltungen)") %>%
  select(-name, -id)
```

```{r echo=FALSE, include=FALSE}
# Votes from abroad are excluded.
# covid_nov %>%
#   filter(str_detect(mun_name, fixed("Ausland")) | str_detect(mun_name, fixed("l'étranger"))) %>%
#   select(mun_name, jaStimmenAbsolut, jaStimmenInProzent)

covid_nov %<>%
  filter(!str_detect(mun_name, fixed("Ausland"))) %>%
  filter(!str_detect(mun_name, fixed("l'étranger"))) %>%
  mutate(GMDNR = as.numeric(mun_id)) %>%
  relocate(GMDNR)
```

Dataset consists of `r scales::number(length(unique(covid_nov$mun_id)), big.mark = ",")` communities.

### Results preview

```{r echo=FALSE}
plot_nationalvotes(
  votedate = "2021-11-28",
  vote_id = 6500, geolevel = "municipality"
)
```

### Link to municipalities

Using `GMDNR` / `mun_id` for deterministic linkage.  

```{r}
covid_nov_gem <- st_gg22 %>%
  left_join(covid_nov)
```

There are again few communities that exist on the map but do not exist in the voting dataset:  

```{r echo=FALSE}
covid_nov_gem %>%
  st_drop_geometry() %>%
  filter(is.na(jaStimmenInProzent)) %>%
  select(GMDNR, GMDNAME)
```

```{r include=FALSE}
write_rds(covid_nov, "data/voting/covid_nov.Rds")
write_rds(covid_nov_gem, "data/voting/covid_nov_gem.Rds")
```
