---
title: "Spatial analyses of 2020-21 excess mortality in CH"
subtitle: "Grid & grided data preparation"
pagetitle: "Spatial Excess: grid"
author: "Radek Panczak & Garyfallos Konstantinoudis"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
    code_folding: show
    self_contained: true
    highlight: pygments
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../docs") })
---

<!-- ----------------------------------------------------- -->

```{r setup, include = FALSE}
set.seed(12345)
options(scipen = 999)
options(max.print = "75")

library(pacman)
p_load(tidyverse, lubridate, magrittr, scales, haven,  
       DT, 
       sf, tmap)

tmap_mode("view")

import::from("sjmisc", "frq")
```

```{r conflicts, include = FALSE}
tidyverse::tidyverse_conflicts()

conflicted::conflict_scout()
```

```{r knit-setup, include = FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

knitr::opts_chunk$set(cache = FALSE,
                      prompt = FALSE,
                      tidy = FALSE,
                      comment = NA,
                      message = FALSE,
                      warning = FALSE)

knitr::opts_knit$set(width = 75)

mu <- Hmisc::markupSpecs$html
```

<!-- ----------------------------------------------------- -->

# Grid

## Municipality boundaries

Used to 'clip' the data to the country outline, 'remove' the area of lakes and introduce the boundaries within grid cells belonging to two different municipalities.   

```{r}
st_gg21 <- read_rds("data/swisstopo/st_gg21.Rds") %>% 
  st_transform(2056)

st_se21_sel <- read_rds("data/swisstopo/st_se21_sel.Rds") %>% 
  st_transform(2056)
```

## Offset

<!-- Openly available STATPOP dataset is used to define `offset` (ie. starting point; lowest left corner) for `st_make_grid`.   -->

```{r eval=FALSE, include=FALSE}
origin_statpop <- read_delim("data-raw/BfS/ag-b-00.03-vz2020statpop/STATPOP2020.zip", 
                             delim = ";", escape_double = FALSE, trim_ws = TRUE) %>% 
  mutate_all(as.integer) %>% 
  summarise(minX = min(E_KOORD),
            minY = min(N_KOORD)) %>% 
  st_as_sf(coords = c("minX", "minY"), 
           crs = 2056,
           remove = FALSE)

qtm(origin_statpop)
```

```{r eval=FALSE, include=FALSE}
offset_statpop <- read_delim("data-raw/BfS/ag-b-00.03-vz2020statpop/STATPOP2020.zip", 
                             delim = ";", escape_double = FALSE, trim_ws = TRUE) %>% 
  mutate_all(as.integer) %>% 
  summarise(minX = min(E_KOORD),
            minY = min(N_KOORD))
```

One of the meteoswiss datasets is used to define `offset` (ie. starting point; lowest left corner) for `st_make_grid`.  

Using meteoswiss origin allows us to perfectly align gridded mortality and population data to meteo data.   

```{r}
p_load(ncdf4)

TabsD_14 <- nc_open('data-raw/meteoswiss/TabsD_ch01r.swiss.lv95_201401010000_201412310000.nc')

E <- ncvar_get(TabsD_14, "E") # X, lon
N <- ncvar_get(TabsD_14, "N") # Y, lat

offset_meteoswiss <- as_tibble(cbind(minX = as.integer(min(E)), 
                                     minY = as.integer(min(N))))

p_unload(ncdf4)
```

```{r eval=FALSE, include=FALSE}
# offset_statpop
offset_meteoswiss

origin_meteoswiss <- st_as_sf(tibble(E = min(E), N = min(N)), 
                              coords = c("E", "N")) %>%
  st_set_crs(2056) 

qtm(origin)
```

## Grid 

1-km grid, with lower left corner defined using `STATPOP` cells.

Grid is then overlaid with *swisstopo municipality data* to select only cells that overlay it.  

At next step the areas of lakes are 'clipped' from the grid.  

Lastly, grid integrating municipality boundaries is also created.

Also, adding ID for each cell (based on row number) and area for each of the iterations.  

```{r eval=FALSE}
grid_1km <- st_gg21 %>% 
  st_make_grid(cellsize = 1000, square = TRUE, 
               offset = c(offset_meteoswiss$minX, offset_meteoswiss$minY)) %>% 
  st_sf() %>%
  st_cast("POLYGON") %>% 
  st_filter(st_gg21, join = st_covers)  %>%
  mutate(ID = row_number()) %>% 
  relocate(ID) %>% 
  st_transform(2056)

# this one takes some time
grid_1km_crop <- grid_1km %>% 
  st_intersection(st_union(st_gg21)) %>% 
  st_collection_extract("POLYGON") %>% 
  mutate(area = st_area(.)) %>% 
  relocate(ID, area) %>%   
  st_transform(2056)

grid_1km_over <- grid_1km_crop %>%
  st_intersection(st_gg21)  %>%
  mutate(ID2 = row_number(),
         area2 = st_area(.)) %>%
  relocate(ID, ID2, area, area2) %>% 
  st_transform(2056)
```

```{r eval=FALSE, include=FALSE}
View(st_drop_geometry(grid_1km))
plot(st_geometry(grid_1km))

View(st_drop_geometry(grid_1km_crop))
plot(st_geometry(grid_1km_crop))

View(st_drop_geometry(grid_1km_over))
plot(st_geometry(grid_1km_over))
```

```{r eval=FALSE, include=FALSE}
# grids
st_write(grid_1km, "data/grid/grid_1km.gpkg", 
         layer = "grid_1km", delete_dsn = TRUE)
st_write(grid_1km_crop, "data/grid/grid_1km.gpkg",
         layer = "grid_1km_crop", append = TRUE)
st_write(grid_1km_over, "data/grid/grid_1km.gpkg", 
         layer = "grid_1km_over", append = TRUE)

# admin bound
st_write(st_gg21, "data/grid/grid_1km.gpkg", 
         layer = "st_gg21", append = TRUE)
st_write(st_se21_sel, "data/grid/grid_1km.gpkg", 
         layer = "st_se21_sel", append = TRUE)

write_rds(grid_1km, "data/grid/grid_1km.Rds")
write_rds(grid_1km_crop, "data/grid/grid_1km_crop.Rds")
write_rds(grid_1km_over, "data/grid/grid_1km_over.Rds")
```

```{r include=FALSE}
grid_1km <- read_rds("data/grid/grid_1km.Rds")
grid_1km_crop <- read_rds("data/grid/grid_1km_crop.Rds")
grid_1km_over <- read_rds("data/grid/grid_1km_over.Rds")
```

```{r eval=FALSE, include=FALSE}
View(st_drop_geometry(grid_1km_over))

grid_1km_over %>% 
  filter(ID == 15) %>% 
  qtm()
```

Example of Bern  

```{r echo=FALSE}
tm_shape(st_gg21 %>% filter(GMDNR == 351)) + 
  tm_borders(col = "darkorchid") +
  tm_shape(st_join(grid_1km,
                   st_gg21 %>% filter(GMDNR == 351),
                   left = FALSE)) + 
  tm_borders(col = "red")
```

## Grid comparison

Example of different integration of boundaries of `Gottlieben` municipality.  

Comparing grid of 'regular' cells (green) and cells 'cropped' by outlines of the country and water bodies. Only grid cells that overlay area of municipality are shown (purple):  

```{r echo=FALSE}
tm_shape(st_gg21 %>% filter(GMDNR == 4651)) + 
  tm_borders(col = "darkorchid") +
  tm_shape(st_join(grid_1km,
                   st_gg21 %>% filter(GMDNR == 4651),
                   left = FALSE)) + 
  tm_borders(col = "forestgreen") +
  tm_shape(st_join(grid_1km_crop,
                   st_gg21 %>% filter(GMDNR == 4651),
                   left = FALSE)) + 
  tm_borders(col = "red")
```

Comparing grid of 'cropped' cells (red again) with grid integrating community boundaries (yellow). Note subdivision of regular cells into smaller ([sometimes very small!](https://github.com/RPanczak/ISPM_geo-mortality/issues/49)) units:   

```{r echo=FALSE}
tm_shape(st_join(grid_1km_crop,
                 st_gg21 %>% filter(GMDNR == 4651),
                 left = FALSE)) + 
  tm_borders(col = "red") +
  tm_shape(st_join(grid_1km_over,
                   st_gg21 %>% filter(GMDNR == 4651),
                   left = FALSE)) + 
  tm_borders(col = "yellow")
```

## Grid centroids

```{r}
grid_1km_cent <- st_centroid(grid_1km)

grid_1km_cent %<>% 
  mutate(X = st_coordinates(.)[, 1],
         Y = st_coordinates(.)[, 2]) %>% 
  relocate(geometry, .after = last_col())
```

```{r include=FALSE}
grid_1km_crop_cent <- st_centroid(grid_1km_crop)

grid_1km_crop_cent %<>% 
  mutate(X = st_coordinates(grid_1km_crop_cent)[, 1],
         Y = st_coordinates(grid_1km_crop_cent)[, 2]) %>% 
  relocate(geometry, .after = last_col())

st_write(grid_1km_cent, "data/grid/grid_1km.gpkg", 
         layer = "grid_1km_cent", append = TRUE)
st_write(grid_1km_crop_cent, "data/grid/grid_1km.gpkg", 
         layer = "grid_1km_crop_cent", append = TRUE)

write_rds(grid_1km_cent, "data/grid/grid_1km_cent.Rds")
write_rds(grid_1km_crop_cent, "data/grid/grid_1km_crop_cent.Rds")
```

```{r eval=FALSE, include=FALSE}
View(st_drop_geometry(grid_1km))
plot(st_geometry(grid_1km))
```

Example of Bern:    

```{r echo=FALSE}
tm_shape(st_gg21 %>% filter(GMDNR == 351)) + 
  tm_borders(col = "darkorchid") +
  tm_shape(st_join(grid_1km,
                   st_gg21 %>% filter(GMDNR == 351),
                   left = FALSE)) + 
  tm_borders(col = "red") +
  tm_shape(st_join(grid_1km_cent,
                   st_gg21 %>% filter(GMDNR == 351),
                   left = FALSE)) + 
  tm_dots() 
```

Of course different grids will result in slightly different centroids in areas that were modified:  

```{r echo=FALSE}
tm_shape(st_join(grid_1km,
                 st_gg21 %>% filter(GMDNR == 4651),
                 left = FALSE)) + 
  tm_borders(col = "red") +
  tm_shape(st_join(grid_1km_crop,
                   st_gg21 %>% filter(GMDNR == 4651),
                   left = FALSE)) + 
  tm_borders(col = "forestgreen") +
  tm_shape(st_join(grid_1km_cent,
                   st_join(grid_1km,
                           st_gg21 %>% filter(GMDNR == 4651),
                           left = FALSE),
                   left = FALSE)) + 
  tm_dots(col = "red") + 
  tm_shape(st_join(grid_1km_crop_cent,
                   st_join(grid_1km,
                           st_gg21 %>% filter(GMDNR == 4651),
                           left = FALSE),
                   left = FALSE)) + 
  tm_dots(col = "forestgreen") 
```

```{r include=FALSE}
rm(grid_1km_crop, grid_1km_crop_cent); gc()
```

<!-- ----------------------------------------------------- -->

# Gridded temperature data

Using gridded data from [meteoswiss](https://www.meteoswiss.admin.ch/home/climate/swiss-climate-in-detail/raeumliche-klimaanalysen.html).  

Using `TabsM`: *Monthly mean temperature* dataset for the years of interest.  

## Monthly data 

Example of two months in 2014:  

```{r echo=FALSE}
p_load(raster)

filename <- "data-raw/meteoswiss/TabsM_ch01r.swiss.lv95_201401010000_201412010000.nc"

# r <- raster::stack(filename) # same same
r <- raster::brick(filename)
raster::crs(r) <- st_crs(2056)$proj4string

# nlayers(r)

plot(r[[ c(1, 7) ]], 
     main = "Monthly temps 2014 - January & July",
     zlim = c(-20, 25), 
     col = rev(RColorBrewer::brewer.pal(11, "RdBu")))

rm(filename, r); gc()
```

```{r eval=FALSE, include=FALSE}
plot(r[[1]], 
     main = "TabsM_14 - January",
     zlim = c(-20, 25), 
     col = rev(RColorBrewer::brewer.pal(11, "RdBu")))

plot(st_geometry(grid_1km_cent), add = TRUE)
```

## Extract

```{r}
extract_temps <- function(filename, year) {
  
  # r <- raster::stack(filename) # same same
  r <- raster::brick(filename)
  raster::crs(r) <- sf::st_crs(2056)$proj4string
  
  ex <- raster::extract(r, grid_1km_cent, df = TRUE) 
  ex <- tibble::as_tibble(ex)
  repeats <- nrow(ex)
  
  ex <- pivot_longer(ex,
                     !ID,
                     names_to = "date_temp",
                     values_to = "temperature"
  )
  
  ex$year <- as.integer(year)
  ex$month <- rep(seq(1, 12), repeats)
  ex$date_temp <- NULL
  
  return(ex)
  
}

temperature <- bind_rows(
  extract_temps("data-raw/meteoswiss/TabsM_ch01r.swiss.lv95_201401010000_201412010000.nc", 2014),
  extract_temps("data-raw/meteoswiss/TabsM_ch01r.swiss.lv95_201501010000_201512010000.nc", 2015),
  extract_temps("data-raw/meteoswiss/TabsM_ch01r.swiss.lv95_201601010000_201612010000.nc", 2016),
  extract_temps("data-raw/meteoswiss/TabsM_ch01r.swiss.lv95_201701010000_201712010000.nc", 2017),
  extract_temps("data-raw/meteoswiss/TabsM_ch01r.swiss.lv95_201801010000_201812010000.nc", 2018),
  extract_temps("data-raw/meteoswiss/TabsM_ch01r.swiss.lv95_201901010000_201912010000.nc", 2019),
  extract_temps("data-raw/meteoswiss/TabsM_ch01r.swiss.lv95_202001010000_202012010000.nc", 2020),
  extract_temps("data-raw/meteoswiss/TabsM_ch01r.swiss.lv95_202101010000_202112010000.nc", 2021)
) %>% 
  filter(! (year == 2021 & month > 6))
```

```{r include=FALSE}
p_unload(raster)
write_rds(temperature, "data/meteoswiss/temperature.Rds")
```

## Examples of time series

Data for Lugano & Jungfraujoch.  

```{r echo=FALSE}
temp_checks <- 
  tibble(Place = c("Jungfraujoch", "Lugano"),
         lon = c(7.980556, 8.9525),
         lat = c(46.547222, 46.005)) %>% 
  st_as_sf(coords = c("lon", "lat"), 
           crs = 4326 , 
           remove = FALSE) %>% 
  st_transform(2056)

# qtm(temp_checks)

temp_checks %<>% 
  st_join(grid_1km) %>% 
  st_drop_geometry() %>% 
  select(Place, ID) %>% 
  left_join(temperature) %>% 
  select(-ID) %>% 
  mutate(Date = ym(paste(year, month, sep = " ")))

ggplot(temp_checks, aes(Date, temperature, 
                        group = Place, col = Place)) +
  geom_line() + 
  geom_point() + 
  theme_light() +
  ylab("Temperature")

rm(temp_checks); gc()
```

<!-- ----------------------------------------------------- -->

# Gridded population data

## Population

```{r, eval=FALSE}
source("R/geom_001_pe_to_grid.R")
grid_1km_over <- read_rds("data/grid/grid_1km_over.Rds")

gc()
r19_pe_grid = geom_001_pe_to_grid("data/STATPOP/r19_pe.dta")
write_rds(r19_pe_grid, "data/STATPOP/r19_pe_grid.Rds")
rm(r19_pe_grid)
gc()

r18_pe_grid = geom_001_pe_to_grid("data/STATPOP/r18_pe.dta")
write_rds(r18_pe_grid, "data/STATPOP/r18_pe_grid.Rds")
rm(r18_pe_grid)
gc()
```

```{r}
pe_grid = bind_rows(read_rds("data/STATPOP/r19_pe_grid.Rds") %>% 
                      st_drop_geometry(),
                    read_rds("data/STATPOP/r18_pe_grid.Rds") %>% 
                      st_drop_geometry())
```

## Aggregate by grid/gemeinde

```{r}
lu_grid_gem = pe_grid %>% 
  group_by(year,sex,age,ID,ID2,area,area2,GMDNR,GMDNAME,KTNR,KTNAME) %>% 
  summarise(pop=sum(n),.groups="drop") %>% 
  dplyr::arrange(year,ID,ID2,sex,age)
rm(pe_grid)
gc()
```

## Expand grid for combinations with 0

```{r}
lu_grid_gem = grid_1km_over %>% 
  st_drop_geometry() %>% 
  as_tibble() %>% 
  dplyr::select(ID,ID2,area,area2,GMDNR,GMDNAME,KTNR,KTNAME) %>% 
  tidyr::expand_grid(year=unique(lu_grid_gem$year),
                     age=unique(lu_grid_gem$age),
                     sex=unique(lu_grid_gem$sex)) %>% 
  dplyr::arrange(year,ID,ID2,sex,age) %>% 
  dplyr::left_join(lu_grid_gem) %>%
  dplyr::mutate(pop=if_else(is.na(pop),as.integer(0),pop))

# filter(lu_grid_gem,GMDNR==4651,age=="<40",sex=="Female",year==2018) %>% view() 
  
```

```{r eval=FALSE}
gc()
write_rds(lu_grid_gem, file="data/blob/lu_grid_gem.rds")
```


## Aggregate by grid

```{r}
st_grid_pop = lu_grid_gem %>% 
  dplyr::group_by(ID,year,age,sex) %>% 
  dplyr::summarise(pop=sum(pop),.groups="drop")
```

```{r eval=FALSE}
gc()
write_rds(st_grid_pop, file="data/blob/st_grid_pop.rds")
```

## Extrapolate pop by month

```{r eval=FALSE}
st_grid_pop_pred = st_grid_pop %>% 
  tidyr::expand_grid(month=1:12) %>% 
  dplyr::mutate(year_month=year+month/12) %>% 
  dplyr::mutate(pop=if_else(month==12,pop,NA_integer_))

# TODO: predict population by month 2014-2021

# m1 = glm(pop ~ age + sex + year_month, data=st_grid_pop_pred, family = poisson(link="log"))
# m2 = lme4::glmer(pop ~ age + sex + year_month + (1|ID), data=st_grid_pop_pred, family = poisson(link="log"))
# m3 = lme4::glmer(pop ~ age + sex + year_month + (year_month|ID), data=st_grid_pop_pred, family = poisson(link="log"))
```

```{r eval=FALSE}
gc()
write_rds(st_grid_pop_pred, file="data/blob/st_grid_pop_pred.rds")
```


# Gridded deaths 2014-2019

### SNC data 

```{r}
snc5_deaths <- read_dta("data/SNC/deaths.dta") %>% 
  zap_formats() %>% zap_labels() %>% zap_label() %>% 
  dplyr::select(-geoyear, -last_census_seen, -dob) %>% 
  dplyr::rename(year = yod) %>% 
  mutate(yod = as.integer(year),
         sex = if_else(sex == "1", "Female", "Male"),
         sex = factor(sex),
         age = cut(age_cont, 
                   breaks = c(0, 40, seq(50, 80, 10), 120), right = FALSE, 
                   labels = c("<40", "40-49", "50-59", "60-69", "70-79", "80+")),
         month = month(dod)) %>% 
  relocate(month, .after = year) %>% 
  dplyr::select(-age_cont, -dod) %>% 
  relocate(age, .after = sex) %>% 
  st_as_sf(coords = c("geox", "geoy"), 
           crs = 2056,
           remove = TRUE)
```

```{r eval=FALSE, include=FALSE}
snc5_deaths %>%
  st_drop_geometry() %>% 
  group_by(age) %>%
  summarise(min = min(year),
            max = max(year))

snc5_deaths %>%
  sample_n(100) %>%
  qtm()

View(st_drop_geometry(snc5_deaths))
```

### Aggregate deaths by ID until 2019

```{r eval=FALSE}
temp <- st_join(snc5_deaths, 
                grid_1km, 
                join = st_intersects) %>% 
  st_drop_geometry() %>% 
  group_by(ID, year, month, sex, age) %>% 
  summarize(deaths = as.integer(n())) %>% 
  ungroup() 

length(unique(grid_1km$ID)) * 
  length(unique(temp$year)) * 
  length(unique(temp$month)) * 
  length(levels(temp$age)) * 
  length(levels(temp$sex))

st_grid_deaths_2014_2019 <-  
  expand_grid(ID = unique(grid_1km$ID), 
              year = unique(temp$year),
              month = unique(temp$month),
              sex = unique(temp$sex),
              age = unique(temp$age)) %>% 
  dplyr::arrange(ID, year, month, sex, age) %>% 
  left_join(temp) %>% 
  mutate(deaths = replace_na(deaths, 0)) %>% 
  mutate(deaths = as.integer(deaths))
```

```{r eval=FALSE}
rm(temp, snc5_deaths); gc()
write_rds(st_grid_deaths_2014_2019, file="data/blob/st_grid_deaths_2014_2019.rds")
```


# Gemeinde level deaths for 2020-2021 into grid

```{r}
st_gem_deaths_2020_2021 =
  read_rds("data/BfS-closed/w_deaths_2015_2021_pop.Rds") %>% 
  dplyr::select(GMDNR,GMDNAME,year,month,age,sex,deaths) %>% 
  dplyr::distinct() %>% 
  dplyr::filter(year >=2020) 
```


```{r}
source("R/geom_002_deaths_to_grid.R")
st_grid_deaths_2020_2021 = geom_002_deaths_to_grid(deaths_data=st_gem_deaths_2020_2021,
                                                   pop_data=lu_grid_gem,
                                                   n_iter=50) # choose number of replications
```


# Join all

```{r eval=FALSE}
st_grid_final = st_grid_pop %>%
  left_join(st_grid_deaths_2014_2019) %>% 
  left_join(st_grid_deaths_2020_2021)

```

